---
author: "Equipo EDUMER"
bibliography: "../input/bib/merit-factorial.bib"
csl: "../input/bib/apa6.csl"
---

```{r}
#| include: false

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse,
               sjmisc, 
               here,
               corrplot,
               ggdist,
               patchwork,
               sjlabelled,
               RColorBrewer)


options(scipen=999)
rm(list = ls())
```

```{r}
#| include: false
#| echo: false

load(file = here("output", "data", "db_long_proc.RData"))

```


# Results


## Descriptive statistics

The distributions of responses for the meritocracy scale are shown in  @fig-likerplot1, distinguishing between perceptions (Panel A) and preferences (Panel B). In Wave 1, Panel A (perceptions of how meritocracy actually operates in society) reveals a strong concentration of agreement in all items, particularly regarding the influence of contacts (77.2%), talent (73.4%), and having rich parents (71.4%), while effort also receives considerable endorsement (67.1%). Nonetheless, a significant proportion of students expressed disagreement: 33.1% did not perceive effort as rewarded, 28.6% rejected the idea that having rich parents leads to success, and 26.6% disagreed that talent plays a decisive role. These results suggest a dual recognition: students acknowledge the strong presence of non-meritocratic factors (contacts and parental wealth) in shaping success, while within the meritocratic dimensions, talent is perceived as more influential than effort. Notably, effort is the dimension perceived as the one with a lowest reward.


```{r}
#| label: fig-likerplot1
#| fig-align: center
#| fig-cap: "Distribution of responses on the scale of perceptions and preferences regarding meritocracy for the first wave"
#| fig-cap-location: top
#| fig-width: 7
#| out-width: '80%'
#| echo: false
#| warning: false
#| message: false

db_likert1 <- db_long %>%
filter(ola == 1) %>% # quedarse con solo la ola 1
select(-c(ola, curse_level, p20, p21_ano, age)) %>% 
na.omit() %>% 
mutate(
  across(
    .cols = c(starts_with(c("perc", "pref"))),
    .fns = ~ sjlabelled::set_labels(., labels = c("Strongly disagree" = 1, "Disagree" = 2, "Agree" = 3, "Strongly agree" = 4))
  )
)

theme_set(theme_ggdist())
colors <- RColorBrewer::brewer.pal(n = 4, name = "RdBu")

a <- db_likert1 %>% 
  select(starts_with("perc")) %>% 
  sjPlot::plot_likert(geom.colors = colors,
                      title = c("A. Perceptions"),
                      geom.size = 0.8,
                      axis.labels = c("Effort", "Talent", "Rich parents", "Contacts"),
                      catcount = 4,
                      values  =  "sum.outside",
                      reverse.colors = F,
                      reverse.scale = T,
                      show.n = FALSE,
                      show.prc.sign = T
                      ) +
  ggplot2::theme(legend.position = "none")

b <- db_likert1 %>% 
  select(starts_with("pref")) %>% 
  sjPlot::plot_likert(geom.colors = colors,
                      title = c("B. Preferences"),
                      geom.size = 0.8,
                     axis.labels = c("Effort", "Talent", "Rich parents", "Contacts"),
                      catcount = 4,
                      values  =  "sum.outside",
                      reverse.colors = F,
                      reverse.scale = T,
                      show.n = FALSE,
                      show.prc.sign = T
  ) +
  ggplot2::theme(legend.position = "bottom")

likerplot <- a / b + plot_annotation(caption = paste0("Source: Own elaboration based on the EDUMER Wave 1 panel survey"," (n = ",dim(db_likert1)[1],")"
))

likerplot
```


In contrast, Panel B (preferences for how meritocracy should operate) displays a reversed pattern. A very large majority (87.9%) endorse effort as the most legitimate basis for reward, far exceeding its perceived importance in practice. Preferences regarding talent are more divided: 54% support rewarding it, while 45% disagree. Attitudes toward non-meritocratic factors are more negative, particularly regarding parental wealth, where responses are split nearly evenly between agreement and disagreement. For contacts, however, a relative majority (60.9%) still expressed that they should be rewarded, compared to 39.1% who disagreed. Overall, these findings highlight a strong normative preference for effort as the central criterion of meritocracy, while talent and parental wealth receive more ambivalent or critical evaluations. The relatively higher acceptance of contacts as a legitimate factor is noteworthy, suggesting that this element may not be viewed as inherently illegitimate in the ideal functioning of society.


```{r}
#| label: fig-likerplot2
#| fig-align: center
#| fig-cap: "Distribution of responses on the scale of perceptions and preferences regarding meritocracy for the second wave"
#| fig-cap-location: top
#| fig-width: 7
#| out-width: '80%'
#| echo: false
#| warning: false
#| message: false

db_likert2 <- db_long %>%
filter(ola == 2) %>% # quedarse con solo la ola 1
select(-c(ola, curse_level, p20, p21_ano, age)) %>% 
na.omit() %>% 
mutate(
  across(
    .cols = c(starts_with(c("perc", "pref"))),
    .fns = ~ sjlabelled::set_labels(., labels = c("Strongly disagree" = 1, "Disagree" = 2, "Agree" = 3, "Strongly agree" = 4))
  )
)

theme_set(theme_ggdist())
colors <- RColorBrewer::brewer.pal(n = 4, name = "RdBu")

a <- db_likert2 %>%
  select(starts_with("perc")) %>% 
  sjPlot::plot_likert(geom.colors = colors,
                      title = c("A. Perceptions"),
                      geom.size = 0.8,
                      axis.labels = c("Effort", "Talent", "Rich parents", "Contacts"),
                      catcount = 4,
                      values  =  "sum.outside",
                      reverse.colors = F,
                      reverse.scale = T,
                      show.n = FALSE,
                      show.prc.sign = T
                      ) +
  ggplot2::theme(legend.position = "none")

b <- db_likert2 %>% 
  select(starts_with("pref")) %>% 
  sjPlot::plot_likert(geom.colors = colors,
                      title = c("B. Preferences"),
                      geom.size = 0.8,
                     axis.labels = c("Effort", "Talent", "Rich parents", "Contacts"),
                      catcount = 4,
                      values  =  "sum.outside",
                      reverse.colors = F,
                      reverse.scale = T,
                      show.n = FALSE,
                      show.prc.sign = T
  ) +
  ggplot2::theme(legend.position = "bottom")

likerplot <- a / b + plot_annotation(caption = paste0("Source: Own elaboration based on the EDUMER Wave 2 panel survey"," (n = ",dim(db_likert2)[1],")"))

likerplot


```


The second wave (@fig-likerplot2) largely replicates these trends, with some shifts. In perceptions (Panel A), the recognition of effort as a determinant of success decreases slightly (61.5%), while the importance attributed to talent (69.9%), parental wealth (69.8%), and contacts (76.6%) remains consistently high. In preferences (Panel B), endorsement of effort as the central criterion of success remains extremely strong (88.2%), showing stability across waves. By contrast, support for talent declines, with disagreement rising to 50.5%. Attitudes toward parental wealth mirror those in Wave 1, remaining evenly divided, while the acceptance of contacts continues to be higher than rejection (59.1% vs. 39.1%), although still below the overwhelming support for effort.

Taken together, both waves reveal a stable duality. Students acknowledge that actual social outcomes are strongly shaped by non-meritocratic factors such as family wealth and contacts, and, to a lesser degree, by talent. At the same time, they overwhelmingly prefer a meritocratic ideal where effort—more than talent or inherited advantage—should determine success. This persistent gap between perceptions of reality and normative preferences underscores a central paradox: students strongly value meritocracy as a principle, while remaining skeptical about its realization in practice.

