---
author: "Equipo EDUMER"
bibliography: "../input/bib/merit-factorial.bib"
csl: "../input/bib/apa6.csl"
---


```{r}
#| echo: false
#| output: false
#| include: false

library(tidyverse)
library(here)

load(file = here("output", "data", "db_long_proc.RData"))


db_long %>% 
  na.omit() %>% 
  group_by(ola) %>% 
  tally()

db_long %>% 
  na.omit() %>% 
  group_by(ola) %>% 
  summarise(m = mean(age),
            sd = sd(age))

db_long %>% 
  na.omit() %>% 
  group_by(ola, p20) %>% 
  tally()
```

# Method

## Participants and Procedure

This study draws on secondary data from the Panel Survey on Education and Meritocracy (EDUMER) for its 2023 and 2024 waves for school-age students. The primary objective is to examine students’ beliefs, attitudes, and behaviors concerning meritocracy, inequality, and citizenship. Data collection was carried out by a professional research firm using a Computer-Assisted Web Interviewing (CAWI) approach, based on online questionnaires administered to sixth-grade and first-year secondary students from nine schools located in the Metropolitan and Valparaíso regions of Chile. The sampling strategy was non-probabilistic and lacked quota controls; however, a minimum sample size of 900 students was established to ensure adequate statistical power. All participants were provided with an informed consent form that was reviewed and validated by a parent or legal guardian. The final analytical sample comprises 846 students for the first wave (386 girls, 421 boys, 39 identifying as other; $M_{age}$ = 13.4, $SD_{age}$ = 1.6), and the second wave followed 662 of them (303 girls, 338 boys, 21 identifying as other; $M_{age}$ = 14.4, $SD_{age}$ = 1.6). 


## Measures

**Scale of Perceptions and Preferences on Meritocracy**: The variables included in the measurement model for meritocratic and non-meritocratic perceptions and preferences are operationalized according to the items proposed in the original scale. Perception of meritocracy is measured by two items that assess the level of agreement with the idea that effort and ability are rewarded in Chile, while non-meritocratic perception is measured by two items evaluating the agreement that success is linked to connections and family wealth. Preference for meritocracy is measured by two items that evaluate agreement with the idea that those who work harder or are more talented should be more rewarded. Preference for non-meritocratic aspects is measured by two indicators assessing agreement that it is acceptable for individuals with better connections or wealthy parents to achieve greater success (see @tbl-merit). Each item is rated on a four-point Likert scale ranging from "strongly disagree" (1) to "strongly agree" (4).   

```{r}
#| tbl-cap: Items for perceptions and preferences for meritocracy scale
#| tbl-cap-location: top
#| echo: false
#| label: tbl-merit

 # remotes::install_github('haozhu233/kableExtra@a6af5c0')  # for problems with collapse_rows()
 pacman::p_load(knitr, kableExtra, dplyr)
 table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }
 tabitems <- read.csv(file = "../input/tables/table-01.csv",header = 1,sep = ",",encoding = "UTF-8") # call file  generated externally for the table
 cnames <- c("Components", "Dimensions","Item (English)" , "Item original (Spanish)")

 knitr::kable(tabitems, table_format, booktabs = T, linesep = "",col.names = cnames) %>%
   kableExtra::kable_styling(
     full_width = T,
     latex_options = c("hold_position"),
     position = "center",
     font_size = 10,
     bootstrap_options=c("striped", "bordered")) %>%
   kableExtra::column_spec(column = 1, width = "1.5cm", ) %>%
   kableExtra::column_spec(column = 2,width = "2 cm") %>%
   kableExtra::collapse_rows(columns = 1:2,valign = "middle")
```

## Analytical strategy

To evaluate the underlying structure of the scale, we employed Confirmatory Factor Analysis (CFA) based on a measurement model with four latent factors, using Diagonally Weighted Least Squares with robust correction (WLSMV) estimation. This estimator is particularly suitable for ordinal data, such as four-point Likert-type scales, as it avoids the bias associated with treating categorical data as continuous [@kline_principles_2023].

Model fit was assessed following the guidelines of Brown [-@brown_confirmatory_2015], using several indices: the Comparative Fit Index (CFI) and the Tucker-Lewis Index (TLI), with acceptable values above 0.95; the Root Mean Square Error of Approximation (RMSEA), with values below 0.06 indicating good fit; and the Chi-square statistic (acceptable fit indicated by $p$ > 0.05 and a Chi-square/df ratio < 3).

A key contribution of this study lies in assessing the metric stability of the measurement model through factorial invariance testing [@davidov_measurement_2014]. We implemented two complementary strategies: invariance across cohorts (primary vs. secondary school students) and longitudinal invariance using two panel waves. Given the ordinal nature of the items, we estimated measurement invariance using methods appropriate for ordered categorical measures, following Liu et al. [-@liu_testing_2017] analysis. Setting a hierarchical approach, we tested four nested models: configural (equal factorial structure), metric (equal factor loadings), scalar (equal intercepts and thresholds), and strict invariance (equal residual variances). 

In addition to the traditional Chi-square difference test for model comparisons, we used changes in fit indices to assess invariance: a change in CFI ($\Delta \geq -0.010$) and RMSEA ($\Delta \geq 0.0$) were adopted as more robust and sample-size-independent criteria, following Chen's recommendations [-@chen_sensitivity_2007].

All analyses were performed using the lavaan package in R version 4.2.2. The hypotheses of this research were [pre-registered](https://osf.io/hjat2/) on the Open Science Framework (OSF), hosted by the Center of Open Science.
