---
bibliography: "../input/bib/merit-factorial.bib"
csl: "../input/bib/apa6.csl"
editor_options: 
  chunk_output_type: console
---


```{r}
#| echo: false
#| output: false
#| include: false

library(tidyverse)
library(here)

load(file = here("output", "data", "db_long_proc.RData"))


db_long %>% 
  na.omit() %>% 
  group_by(ola) %>% 
  tally()

db_long %>% 
  na.omit() %>% 
  group_by(ola) %>% 
  summarise(m = mean(age),
            sd = sd(age))

db_long %>% 
  na.omit() %>% 
  group_by(ola, p20) %>% 
  tally()

db_long %>% 
  na.omit() %>% 
  group_by(cohort_level) %>% 
  tally()

db_long %>% 
  na.omit() %>% 
  group_by(cohort_level) %>% 
  summarise(m = mean(age),
            sd = sd(age))

db_long %>% 
  na.omit() %>% 
  group_by(cohort_level, p20) %>% 
  tally()

```

# Method

## Participants

This study draws on secondary data from the Panel Survey on Education and Meritocracy (EDUMER) for the 2023 and 2024 waves, covering school-age students[^1]. The primary objective of the survey was to examine students’ beliefs, attitudes, and behaviors concerning meritocracy, inequality, and citizenship. The questionnaire was administered to sixth-grade and first-year secondary students from nine schools in the Metropolitan and Valparaíso regions of Chile. The sampling strategy was non-probabilistic and lacked quota controls; however, a minimum sample size of 900 students was established to ensure adequate statistical power. After data processing and the elimination of missing cases (listwise deletion), the final analytical sample comprises 846 students for the first wave (386 girls, 421 boys, 39 identifying as other; $M_{age}$ = 13.4, $SD_{age}$ = 1.6), and the second wave followed 662 of them (303 girls, 338 boys, 21 identifying as other; $M_{age}$ = 14.4, $SD_{age}$ = 1.6), corresponding to a retention rate of 78.3% (attrition 21.7%). The primary causes of attrition were school transfers between academic years and student absenteeism, which was reported to be high in several establishments. 

[^1]: The survey was collected within the research project “Meritocracy in schools: Moral foundations of the education market and its implications for civic education in Chile” (FONDECYT Regular 2021 Nº 1210847).

## Procedure

Data collection was conducted by a professional research firm using a Computer-Assisted Web Interviewing (CAWI) approach, based on online questionnaires. All participants received an informed consent form that was reviewed and validated by a parent or legal guardian before participation. To enable longitudinal linkage while protecting confidentiality, the survey firm generated anonymous unique identifiers for each student; no directly identifying information (e.g., names) was collected for students or their guardians in the analytic files.

## Measures

### Scale of perceptions and preferences of meritocracy

Meritocratic and non-meritocratic perceptions and preferences were measured using the same items proposed in the original scale [@castillo_multidimensional_2023]. Each of the four dimensions of the scale is measured by two items. The dimensions are:

- Perceived meritocracy: the extent to which effort and ability are rewarded in Chile,

- Perceived non-meritocracy: the extent to which success is perceived as linked to connections and family wealth,

- Preference for meritocracy: agreement with that those who work harder or are more talented should be better rewarded,

- Preference for non-meritocracy: agreement with that it is acceptable for individuals with better connections or wealthy parents to achieve greater success  (see @tbl-merit).
 
Each item is rated on a four-point Likert scale ranging from “strongly disagree” (1) to “strongly agree” (4), therefore higher scores indicate stronger endorsement of the corresponding perception or preference.

```{r}
#| tbl-cap: Items for perceptions and preferences for meritocracy scale
#| tbl-cap-location: top
#| echo: false
#| label: tbl-merit

 # remotes::install_github('haozhu233/kableExtra@a6af5c0')  # for problems with collapse_rows()
 pacman::p_load(knitr, kableExtra, dplyr)
 table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }
 tabitems <- read.csv(file = "../input/tables/table-01.csv",header = 1,sep = ",",encoding = "UTF-8") # call file  generated externally for the table
 cnames <- c("Components", "Dimensions","Item (English)" , "Item original (Spanish)")

 knitr::kable(tabitems, "latex", booktabs = T, linesep = "",col.names = cnames) %>%
   kableExtra::kable_styling(
     full_width = T,
     latex_options = c("hold_position"),
     position = "center",
     font_size = 9.5,
     bootstrap_options=c("striped", "bordered")) %>%
   kableExtra::column_spec(column = 1, width = "2cm", ) %>%
   kableExtra::column_spec(column = 2,width = "2 cm") %>%
   kableExtra::collapse_rows(columns = 1:2,valign = "middle")
```



### Cohort level

To distinguish students’ academic level across waves for conditional and multi-group invariance analyses, we created a cohort indicator capturing whether the respondent belonged to the primary (sixth grade) or secondary (first-year secondary) cohort at the time of each survey wave. Descriptive statistics[^2] by cohort are reported in @tbl-cohort. 

[^2]: @tbl-cohort totals (N = 1,508) correspond to the sum of person-wave observations (846 at T1 + 662 at T2), rather than unique individuals.


```{r}
#| tbl-cap: Descriptive statistics for cohort level
#| tbl-cap-location: top
#| echo: false
#| label: tbl-cohort
#| message: false
#| warning: false


tot <- db_long %>% 
  na.omit() %>% 
  group_by(ola, cohort_level) %>% 
  tally() %>% 
  rename(n_total = n) %>% 
  mutate(ola = if_else(ola == "1", "Wave 1", "Wave 2")) %>% 
  ungroup()

tot <- tot %>% 
    add_row(
      ola = "Wave 1",
      cohort_level = "Total",
      n_total = sum(tot[c(1,2),3])
    ) %>% 
  add_row(
      ola = "Wave 2",
      cohort_level = "Total",
      n_total = sum(tot[c(3,4),3]))


edad <- db_long %>% 
  na.omit() %>% 
  group_by(ola, cohort_level) %>% 
  summarise(m = round(mean(age),2),
            sd = round(sd(age),2),
            .groups = "drop_last") %>% 
  mutate(ola = if_else(ola == "1", "Wave 1", "Wave 2")) %>% 
  ungroup()

edad_ola <- db_long %>% 
  na.omit() %>% 
  group_by(ola) %>% 
  summarise(m = round(mean(age),2),
            sd = round(sd(age),2),
            .groups = "drop_last") %>% 
  mutate(ola = if_else(ola == "1", "Wave 1", "Wave 2"),
         cohort_level = "Total") %>% 
  select(ola, cohort_level, m, sd)


edad <- bind_rows(edad, edad_ola)

gen <- db_long %>% 
  na.omit() %>% 
  group_by(ola, cohort_level, p20) %>% 
  tally() %>% 
  ungroup() %>% 
  pivot_wider(id_cols = c(ola, cohort_level),
              names_from = p20,
              values_from = n) %>% 
  mutate(ola = if_else(ola == "1", "Wave 1", "Wave 2"))

gen_ola <- db_long %>% 
  na.omit() %>% 
  group_by(ola, p20) %>% 
  tally() %>% 
  ungroup() %>% 
  pivot_wider(id_cols = c(ola),
              names_from = p20,
              values_from = n) %>% 
  mutate(ola = if_else(ola == "1", "Wave 1", "Wave 2"),
         cohort_level = "Total") %>% 
  select(ola, cohort_level, H, M, O)

gen <- bind_rows(gen, gen_ola)

left_join(tot, edad) %>% 
  left_join(., gen) %>% 
  arrange(ola) %>% 
   knitr::kable(., "latex", escape = T,
                booktabs = T, 
                linesep = "",
                col.names = c("Wave", "Cohort level", "N", "Mean", "SD", "Man", "Women", "Others")) %>%
    kableExtra::add_header_above(c(" " = 3, "Age" = 2, "Gender" = 3)) %>% 
   kableExtra::kable_styling(
     full_width = T,
     latex_options = c("hold_position"),
     position = "center",
     font_size = 9.5,
     bootstrap_options=c("striped", "bordered")) %>% 
  kableExtra::collapse_rows(1)

```




## Analytical strategy

### Measurement model and estimation

To evaluate the underlying structure of the scale, we employed Confirmatory Factor Analysis (CFA) based on a measurement model with four latent factors, using Diagonally Weighted Least Squares with robust correction (WLSMV) estimation. This estimator is particularly suitable for ordinal data, such as four-point Likert-type scales, as it avoids the bias associated with treating categorical data as continuous [@kline_principles_2023].

Model fit was assessed following the guidelines of Brown [-@brown_confirmatory_2015], using several indices: the Comparative Fit Index (CFI) and the Tucker-Lewis Index (TLI), with acceptable values above 0.95; the Root Mean Square Error of Approximation (RMSEA), with values below 0.06 indicating good fit; and the Chi-square statistic (acceptable fit indicated by $p$ > 0.05 and a Chi-square/df ratio < 3).

### Measurement invariance

A key contribution of this study is its assessment of measurement stability through factorial invariance testing [@davidov_measurement_2014]. We implemented two complementary strategies: cross-group invariance across school levels (primary vs. secondary students) and longitudinal invariance across two panel waves. Given the ordinal nature of the items, we used estimation procedures appropriate for ordered categorical indicators, following Liu et al. [-@liu_testing_2017] for longitudinal invariance and @svetina_multiplegroup_2020 for cross-group tests. For the longitudinal analyses, we adopted the standard hierarchical sequence of four nested models—configural (same factor structure), metric (equal loadings), scalar (equal thresholds/intercepts), and strict (equal residual variances). For the cross-group analyses, we estimated a configural model, then a threshold-invariant model, and finally a model jointly constraining thresholds and factor loadings across groups.

In addition to the traditional chi-square difference test, we assessed invariance using changes in approximate fit indices. Following Chen’s [-@chen_sensitivity_2007] recommendations for adequate sample sizes, we treated $\Delta$CFI as the primary criterion and $\Delta$RMSEA as a supplementary criterion. Noninvariance was indicated by a decrease in CFI of at least .010 (i.e., $\Delta$CFI $\leq$ −0.010), together with an increase in RMSEA of at least .015 (i.e., $\Delta$RMSEA $\geq$ 0.015).


All analyses were performed using the *lavaan* package in R version 4.5.2. The hypotheses of this research were [pre-registered](https://osf.io/hjat2/) on the Open Science Framework (OSF).
