---
author: "Equipo EDUMER"
bibliography: "../input/bib/merit-factorial.bib"
csl: "../input/bib/apa6.csl"
---

## Validity evidence: Latent correlations between meritocratic beliefs and market justice preferences

```{r}
#| echo: false
#| message: false
#| warning: false

library(knitr)
library(here)
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      error = F, 
                      message = F) 

table_format <- if (is_html_output()) {
  "html"
} else if (is_latex_output()) {
  "latex"
}
table_format2 <- if (is_html_output()) {
  T
} else if (is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```


```{r}
#| include: false
#| echo: false
#| results: false

source(here("processing/prod_analysis_validez.R"))
```

Market justice preferences were modeled as a single latent factor in each wave, indicated by three ordinal items capturing attitudes toward education, healthcare, and pensions. The one-factor structure showed satisfactory fit, and longitudinal measurement invariance was supported across waves (configural, metric, threshold, and strict). Full invariance results are reported in the Supplementary Material.

To provide construct-validity evidence, we estimated a joint CFA model combining the longitudinally invariant meritocracy factors with the longitudinally invariant market-justice-preferences factor. Given the ordinal nature of the indicators, models were estimated using WLSMV with theta parameterization. As reported in @tbl-fit-validity, the model with freely estimated correlations across waves showed good fit ($\chi^2$(170)=534.699, CFI = 0.940, TLI = 0.918, RMSEA = 0.061). Imposing equality constraints on the corresponding correlations across waves yielded a slightly better-fitting model ($\chi^2$(174)=509.218, CFI = 0.945, TLI = 0.927, RMSEA = 0.058).


```{r}
#| tbl-cap: "Model fit for latent-correlation validity models"
#| tbl-cap-location: top
#| echo: false
#| warning: false
#| label: tbl-fit-validity

get_fit <- function(fit, label){

  fm <- fitMeasures(
    fit,
    c("chisq.scaled","df","pvalue.scaled","cfi.scaled","tli.scaled","rmsea.scaled")
  )

  tibble(
    Model = label,
    `χ²` = round(fm["chisq.scaled"], 3),
    df = as.integer(fm["df"]),
    `p value` = ifelse(fm["pvalue.scaled"] < .001, "<.001", sprintf("%.3f", fm["pvalue.scaled"])),
    `CFI` = round(fm["cfi.scaled"], 3),
    `TLI` = round(fm["tli.scaled"], 3),
    `RMSEA` = round(fm["rmsea.scaled"], 3))
}

fit_tab <- bind_rows(
  get_fit(fit_validity_free,  "Free correlations (by wave)"),
  get_fit(fit_validity_equal, "Equal correlations (stable across waves)")
)

fit_tab %>%
  kableExtra::kable("html",
        caption = NULL) %>%
  kableExtra::kable_styling(latex_options = "hold_position",
                            bootstrap_options=c("striped", "bordered", "condensed"),
                full_width = FALSE) %>% 
  kableExtra::add_footnote("N = 583", notation = "none")
```

The superior fit of the equal-correlations model indicates that the associations between market justice preferences and meritocracy dimensions remained stable across waves, supporting the temporal consistency of these relationships. We next examined the within-wave latent correlations to assess construct validity (see @tbl-latent-cor). The pattern of associations revealed theoretically meaningful distinctions. Market justice preferences showed modest positive correlations with preferred meritocracy in both waves (Wave 1: $r$ = 0.185, p < .05; Wave 2: $r$ = 0.192, p < .05), suggesting that support for market-based allocation aligns with, but is not reducible to, meritocratic ideals. 

More strikingly, market justice preferences exhibited substantially stronger associations with preferred non-meritocratic principles—specifically, endorsement of personal contacts or wealthy family background as legitimate bases for reward—with correlations more than twice as large as those for meritocratic preferences (Wave 1: $r$ = 0.441, p < .05; Wave 2: $r$ = 0.468, p < .05). 

This asymmetric pattern provides both convergent and discriminant validity evidence: while market justice preferences relate positively to meritocratic principles, they are more strongly characterized by acceptance of market-generated inequalities that extend beyond merit, encompassing advantages derived from social networks and inherited wealth. These findings situate market justice preferences within a broader ideological framework that legitimates market-based differentiation through multiple distributive principles, not solely meritocratic ones.


```{r}
#| tbl-cap: Standardized latent correlations between market justice preferences and meritocracy dimensions (within-wave only)"
#| tbl-cap-location: top
#| echo: false
#| warning: false
#| label: tbl-latent-cor

pe1 <- parameterEstimates(fit_validity_free, standardized = TRUE)

mjp_vars   <- c("mjp1","mjp2")
merit_vars <- c("percmerit1","percmerit2","percnmerit1","percnmerit2",
                "prefmerit1","prefmerit2","prefnmerit1","prefnmerit2")

tab_mat <- pe1 %>%
  filter(
    op == "~~",
    (lhs %in% mjp_vars & rhs %in% merit_vars) | (rhs %in% mjp_vars & lhs %in% merit_vars)
  ) %>%
  mutate(
    mjp_raw   = if_else(lhs %in% mjp_vars, lhs, rhs),
    merit_raw = if_else(lhs %in% mjp_vars, rhs, lhs),
    mjp_w     = str_extract(mjp_raw,   "[12]$"),
    merit_w   = str_extract(merit_raw, "[12]$")
  ) %>%
  filter(mjp_w == merit_w) %>%
  mutate(
    mjp_col = if_else(mjp_w == "1", "MJP (Wave 1)", "MJP (Wave 2)"),
    merit   = str_remove(merit_raw, "[12]$"),
    star    = if_else(!is.na(pvalue) & pvalue < .05, "*", ""),
    cell    = paste0(sprintf("%.3f", std.all), star)
  ) %>%
  select(merit, mjp_col, cell) %>%
  distinct() %>%
  pivot_wider(names_from = mjp_col, values_from = cell) %>%
  mutate(merit = recode(
    merit,
    percmerit  = "Meritocratic perception",
    percnmerit = "Non-meritocratic perception",
    prefmerit  = "Meritocratic preferences",
    prefnmerit = "Non-meritocratic preferences",
    .default = merit
  )) %>%
  rename(Construct = merit)
tab_mat %>%
  kableExtra::kable("html", escape = TRUE,
        caption = NULL) %>%
  kableExtra::kable_styling(latex_options = "hold_position",
                            bootstrap_options=c("striped", "bordered", "condensed"), 
                            full_width = FALSE) %>% 
  kableExtra::add_footnote(c("N = 583","*p < .05."), "none")

```

To evaluate the temporal stability of these associations, we compared the model with freely estimated correlations across waves to a model constraining corresponding within-wave correlations to equality. The constrained model demonstrated superior fit (as shown in @tbl-fit-validity), indicating that the equality constraints were supported overall. However, inspection of the point estimates reveals some noteworthy patterns. 

The associations with meritocratic and non-meritocratic preferences remained highly consistent across waves (meritocratic: $r$ = 0.185 to 0.192; non-meritocratic: $r$ = 0.441 to 0.468), reinforcing their robustness. In contrast, correlations with the perception dimensions showed more descriptive variability: the association with meritocratic perceptions weakened from Wave 1 ($r$ = 0.233, p < .05) to Wave 2 ($r$ = 0.126, ns), while the correlation with non-meritocratic perceptions shifted from near-zero ($r$ = 0.006) to negative ($r$ = -0.112, p < .05). 

Despite this variation, the restricted model fit suggests that these differences do not represent a substantially significant change, but rather reflect variability around stable parameters. Overall, the evidence supports the temporal stability of the structural relationships between market justice preferences and dimensions of meritocratic beliefs, particularly with regard to theoretically central preference constructs. Taken together, these validity findings illustrate the fundamental purpose of the meritocracy scale: to provide a multidimensional assessment that distinguishes between what individuals perceive as actual operating principles in society (perceptions) and what they believe should guide distribution (preferences), while differentiating between meritocratic and non-meritocratic criteria within each domain. The differential patterns of association with market justice preferences—in particular, the stronger links with non-meritocratic dimensions—demonstrate that the scale captures theoretically meaningful variation that would be obscured by unidimensional measures.

