---
bibliography: "../input/bib/merit-factorial.bib"
csl: "../input/bib/apa6.csl"
editor_options:
  chunk_output_type: console
execute:
  cache: true
  echo: false
bib: 
  - "../input/bib/merit-factorial.bib"
---

# Results

This section is organized into three parts, each one referred to corresponding three research questions and hypotheses. First, we present descriptive statistics and visualizations of the survey items measuring meritocratic perceptions and preferences, as well as perceived privilege and its acceptance. Second, we report the results of the confirmatory factor analysis (CFA) conducted to evaluate the proposed four-factor structure of meritocratic perceptions, meritocratic preferences, perceived privilege, and privilege acceptance. Based on in these results we analyze the associations between the different dimensions of meritocracy and privilege, which is at the core of our investigation. Finally, we present the results of multigroup analyses to assess measurement invariance across cohorts and waves.

## Descriptives

```{r setup, include=FALSE}
#| echo: false
#| message: false
#| warning: false

library(knitr)

knitr::opts_chunk$set(echo = F,
                      warning = F,
                      error = F, 
                      message = F) 

table_format <- if (is_html_output()) {
  "html"
} else if (is_latex_output()) {
  "latex"
}
table_format2 <- if (is_html_output()) {
  T
} else if (is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")

if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,
  sjmisc,
  sjPlot,
  here,
  lavaan,
  psych,
  corrplot,
  ggdist,
  patchwork,
  semTools,
  gtools,
  kableExtra,
  scales
)

options(scipen = 999)
```

```{r setup-data-db-long, include=FALSE}
#| echo: false
#| output: false

load(file = here("output", "data", "db_long_proc.RData"))

names(db_long)
glimpse(db_long)
dim(db_long)
```


@fig-likerplot1 displays the response distributions for the meritocracy and privilege items by each survey wave, distinguishing perceptions from preferences. A clear and stable gap emerges between how students perceive social rewards to operate and how they think they should operate. In the perception items, agreement is consistently high for contacts, parental wealth, and talent, whereas effort displays the lowest perception.  In the preference items, by contrast, effort is overwhelmingly endorsed as the legitimate basis for rewards, exceeding by far its perceived role in practice. Noticeably, in terms of preferences, is the talent item the one showing the lower frequency, being the use of contacts and family resources more preferred than this meritocratic stance. Roughly, more that half of respondents in both waves express an acceptance of the influence of rich parents and contacts, revealing that such privilege criteria are not uniformly seen as illegitimate way to get ahead. 


```{r fig-likerplot1, echo=FALSE}
#| fig-align: center
#| fig-cap: "Distribution of responses on the scale of perceptions and preferences regarding meritocracy by wave"
#| fig-cap-location: top
#| fig-asp: 0.9
#| out-width: '100%'
#| warning: false
#| message: false

library(cowplot)

cols  <- RColorBrewer::brewer.pal(4, "RdBu")
items1 <- c("perc_effort","perc_talent","perc_rich_parents","perc_contact")

db_likert1 <- db_long %>%
  filter(ola == 1) %>%
  select(starts_with(c("perc"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items1),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert1 <- db_likert1 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab1 <- db_likert1 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

a <- ggplot(db_likert1, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab1, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(title = "Perceptions", x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 4, 5, 10))

items2 <- c("pref_effort","pref_talent","pref_rich_parents","pref_contact")

db_likert2 <- db_long %>%
  filter(ola == 1) %>%
  select(starts_with(c("pref"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items2),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert2 <- db_likert2 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab2 <- db_likert2 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

b <- ggplot(db_likert2, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab2, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(title = "Preferences", x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        plot.margin = margin(5, 18, 5, 0))

# ----


db_likert3 <- db_long %>%
  filter(ola == 2) %>%
  select(starts_with(c("perc"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items1),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert3 <- db_likert3 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab3 <- db_likert3 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

c <- ggplot(db_likert3, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab3, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 4, 5, 10))

db_likert4 <- db_long %>%
  filter(ola == 2) %>%
  select(starts_with(c("pref"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items2),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert4 <- db_likert4 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab4 <- db_likert4 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

d <- ggplot(db_likert4, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab4, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        plot.margin = margin(5, 18, 5, 0))
# 4) grilla 2x2
# sin align horizontal: si no, el panel de preferencias hereda el ancho del eje Y y deja un hueco
panels <- cowplot::plot_grid(a, b, c, d, ncol = 2, rel_widths = c(1.25, 1))

# 5) poner Wave 1 / Wave 2 encima (no consume ancho)
panels_labeled <- cowplot::ggdraw(panels) +
  draw_label("Wave 1", x = 0.03, y = 0.75, angle = 0, hjust = 1.5, vjust = 0.5, size = 11) +
  draw_label("Wave 2", x = 0.03, y = 0.25, angle = 0, hjust = 1.5, vjust = 0.5, size = 11)

# 6) leyenda a mano COMO PANEL (sin extraer grobs)
lik_labels <- c("Strongly disagree", "Disagree", "Agree", "Strongly agree")
df_leg <- tibble(cat = factor(lik_labels, levels = lik_labels))

legend_only <- ggplot(df_leg, aes(x = 2, y = 2, fill = cat)) +
  geom_point(shape = 22, size = 6, color = "black", stroke = 0.3) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), clip = "on") +  # “esconde” los puntos
  scale_fill_manual(values = cols, drop = FALSE) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE, title = NULL)) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.text     = element_text(size = 9),
    legend.key.size = unit(4, "mm"),
    plot.margin     = margin(0, 0, 0, 0)
  )
# 7) caption común como panel
cap <- cowplot::ggdraw() +
  draw_label("Source: own elaboration based on EDUMER Panel Survey (N Wave 1 =  846; N Wave 2 = 662)", x = 0, hjust = 0, size = 9)

# 8) final: todo junto
final <- plot_grid(
  panels_labeled,
  legend_only,
  cap,
  ncol = 1,
  rel_heights = c(1.5, 0.10, 0.06)
)

final +
  theme(plot.margin = margin(t = 5.5, r = 5.5, b = 5.5, l = 50))

```


Regarding the primary and secondary education cohorts, some contrasts emerge as seen in @fig-likerplot2. In terms of perception, the meritocratic items (effort and talent) are the ones with the highest agreement in the youngest cohort and the lowest in the oldest one. At the same time, the privilege perceptual items (contacts and family wealth) are the ones with the lowest agreement in the youngest cohort and the highest in the oldest one. This pattern is consistent with evidence on a "disillusionment" with meritocracy as students progress through the educational system [@fan_disillusionment_2026; @tang_meritocratic_2025]. In terms of preferences, the pattern is more stable across cohorts, with effort being the most endorsed legitimate basis for reward, while talent, family wealth, and especially contacts remain more ambivalent. As in the previous descriprive analysis by wave, contacts are more often accepted than rejected in both cohorts. Overall, the cohort comparison reveals a widening gap between an effort-centered meritocratic ideal and the perceived importance of privilege advantages in practice.    

```{r fig-likerplot2, echo=FALSE}
#| fig-align: center
#| fig-cap: "Distribution of responses on the scale of perceptions and preferences regarding meritocracy by cohort"
#| fig-cap-location: top
#| fig-asp: 0.9
#| out-width: '100%'
#| warning: false
#| message: false
cols  <- RColorBrewer::brewer.pal(4, "RdBu")
items1 <- c("perc_effort","perc_talent","perc_rich_parents","perc_contact")

db_likert1 <- db_long %>%
  filter(cohort_level == "Primary") %>%
  select(starts_with(c("perc"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items1),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert1 <- db_likert1 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab1 <- db_likert1 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

a <- ggplot(db_likert1, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab1, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(title = "Perceptions", x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 4, 5, 10))

items2 <- c("pref_effort","pref_talent","pref_rich_parents","pref_contact")

db_likert2 <- db_long %>%
  filter(cohort_level == "Primary") %>%
  select(starts_with(c("pref"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items2),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert2 <- db_likert2 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab2 <- db_likert2 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

b <- ggplot(db_likert2, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab2, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(title = "Preferences", x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        plot.margin = margin(5, 18, 5, 0))

# ----


db_likert3 <- db_long %>%
  filter(cohort_level == "Secondary") %>%
  select(starts_with(c("perc"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items1),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert3 <- db_likert3 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab3 <- db_likert3 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

c <- ggplot(db_likert3, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab3, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 4, 5, 10))

db_likert4 <- db_long %>%
  filter(cohort_level == "Secondary") %>%
  select(starts_with(c("pref"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items2),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert4 <- db_likert4 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab4 <- db_likert4 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

d <- ggplot(db_likert4, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab4, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        plot.margin = margin(5, 18, 5, 0))
# 4) grilla 2x2
panels <- plot_grid(a, b, c, d, ncol = 2, rel_widths = c(1.25, 1))

# 5) poner Wave 1 / Wave 2 encima (no consume ancho)
panels_labeled <- cowplot::ggdraw(panels) +
  draw_label("Primary", x = 0.03, y = 0.75, angle = 0, hjust = 1.2, vjust = 0.5, size = 11) +
  draw_label("Secondary", x = 0.03, y = 0.25, angle = 0, hjust = 1.2, vjust = 0.5, size = 11)

# 6) leyenda a mano COMO PANEL (sin extraer grobs)
lik_labels <- c("Strongly disagree", "Disagree", "Agree", "Strongly agree")
df_leg <- tibble(cat = factor(lik_labels, levels = lik_labels))

legend_only <- ggplot(df_leg, aes(x = 2, y = 2, fill = cat)) +
  geom_point(shape = 22, size = 6, color = "black", stroke = 0.3) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), clip = "on") +  # “esconde” los puntos
  scale_fill_manual(values = cols, drop = FALSE) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE, title = NULL)) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.text     = element_text(size = 9),
    legend.key.size = unit(4, "mm"),
    plot.margin     = margin(0, 0, 0, 0)
  )
# 7) caption común como panel
cap <- cowplot::ggdraw() +
  draw_label("Source: own elaboration based on EDUMER Panel Survey (N Primary =  725; N Secondary = 783)", x = 0, hjust = 0, size = 9)

# 8) final: todo junto
final <- plot_grid(
  panels_labeled,
  legend_only,
  cap,
  ncol = 1,
  rel_heights = c(1.5, 0.10, 0.06)
)

final +
  theme(plot.margin = margin(t = 5.5, r = 5.5, b = 5.5, l = 50))

```

## Multidimensional measurement of meritocracy and privilege

```{r setup-measurement}
#| echo: false
#| message: false
#| warning: false
#| results: false

library(knitr)

knitr::opts_chunk$set(echo = F,
                      warning = F,
                      error = F, 
                      message = F) 

table_format <- if (is_html_output()) {
  "html"
} else if (is_latex_output()) {
  "latex"
}
table_format2 <- if (is_html_output()) {
  T
} else if (is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")

if (!require("pacman")) install.packages("pacman") #para instalar pacman en caso de no estar instalado

pacman::p_load(
  tidyverse,
  sjmisc,
  sjPlot,
  here,
  lavaan,
  psych,
  ggdist,
  semTools,
  gtools,
  kableExtra
)

# Silence only the known Heywood warning from lavaan while preserving all others
quiet_heywood <- function(expr) {
  withCallingHandlers(
    expr,
    warning = function(w) {
      if (grepl("some estimated ov variances are negative", conditionMessage(w), fixed = TRUE)) {
        invokeRestart("muffleWarning")
      }
    }
  )
}
run_cfa <- function(...) quiet_heywood(cfa(...))

options(scipen = 999) # para eliminar la notación científica

load(file = here("output", "data", "db_long_proc.RData")) #cargar base con las dos olas 

names(db_long) #ver nombres de variables
glimpse(db_long) # ver los primeros 15 casos de cada variable
dim(db_long) # ver cantidad de filas y columnas 
```

```{r generate-data}
#| echo: false
#| results: false

db1 <- db_long %>%
select(id_estudiante, starts_with(c("perc", "pref"))) %>% 
  drop_na()#%>% # eliminar variables que no son parte del analisis
#na.omit() # listwise deletion

```


```{r pooled-model}
#| include: false 

model_base <- ('
perc_merit =~ perc_effort + perc_talent
perc_nmerit =~ perc_rich_parents + perc_contact
pref_merit =~ pref_effort + pref_talent
pref_nmerit =~ pref_rich_parents + pref_contact
')

mgeneral_cfa <- run_cfa(model = model_base, 
                   data = db1, 
                   estimator = "WLSMV",
                   ordered = T,
                   std.lv = F,
                   parameterization = "theta")

summary(mgeneral_cfa, standardized = TRUE, fit.measures = TRUE)
```

The eitgh items described in the previous section are conceptually organized into four latent dimensions: meritocratic perceptions, meritocratic preferences, perceived privilege, and privilege acceptance. Previous research in adult population has shown that these dimensions are empirically distinguishable and can be measured with a four-factor structure [@castillo_multidimensional_2023]. In this study, we extend this line of research to the adolescent population, examining whether the same four-factor structure holds in our sample of students from primary and secondary education cohorts ($H_1$).

@fig-general presents a diagram of the general measurement model. The hypothesized four latent dimensions are represented by circles in the middle, each correspondig to two observed indicators (rectangles) that are expected to load on the corresponding latent factor. The arrows indicate the direction of the hypothesized relationships, with factor loadings representing the strength of the association between each indicator and its latent construct. The model also allows for correlations among the four latent factors, reflecting the potential interrelationships between meritocratic perceptions, meritocratic preferences, perceived privilege, and privilege acceptance. The model shows a reasonable --though not fully satisfactory-- fit to the data (CFI = 0.97; TLI = 0.94; RMSEA = 0.07). Although the CFI falls within commonly accepted ranges, the RMSEA and TLI remain below conventional thresholds, indicating that fit is acceptable in some respects but not uniformly strong across indices. As we will see later on, part of the misfit relates to some differences across cohorts, which are explored in the subsequent multigroup analyses.


```{r}
#| label: fig-general
#| fig-cap: Measurement model diagram (pooled sample)
#| fig-cap-location: top
#| echo: false
#| fig-width: 10
#| fig-height: 5.3
#| out-width: '100%'
#| cache.extra: !expr unname(tools::md5sum(here::here('processing', 'plot_cfa_diagram.R')))

# Path diagram of the pooled 4-factor CFA drawn from a lavaan fit object
source(here::here("processing", "plot_cfa_diagram.R"))
plot_cfa_diagram(mgeneral_cfa)
```


All standardized factor loadings are statistically significant and generally moderate to strong, though their magnitude varies across dimensions. For meritocratic perceptions, the effort item shows a particularly strong loading ($\beta$ = 0.88) and talent a moderately strong loading ($\beta$ = 0.62). Meritocratic preferences exhibit more even, but somewhat weaker, loadings for effort ($\beta$ = 0.59) and talent ($\beta$ = 0.55). For perceived privilege, both indicators load strongly, especially contacts ($\beta$ = 0.88), followed by coming from a rich family ($\beta$ = 0.72). Privilege acceptance also shows strong and relatively balanced loadings, with contacts ($\beta$ = 0.84) and rich family ($\beta$ = 0.72), indicating that both indicators are similarly representative of this latent dimension. In general, the factor loadings suggest that the observed indicators are reasonably good measures of their respective latent constructs, with some variation in strength across dimensions.

The analysis of latent correlations gives us important insights into the relationships among the four dimensions ($H_2$). As stated before, allowing these for dimensions to correlate pertmit us to open the black-box of meritocracy and privilege beliefs, which both in the common sense as well as in the literature are often treated as the oposite poles of a single continuum. The results show that the four dimensions are indeed interrelated, but not perfectly aligned, indicating that students' beliefs about meritocracy and privilege are complex and multifaceted. To begin with, all the correlations but one are actually positive. The only negative correlation is between meritocratic perceptions and privilege perceptions, which is consistent with the idea that perceiving society as more meritocratic is associated with a lower recognition of the role of privilege (and vice versa). Some positive associations also show a more intuitive pattern, such as the positive correlation between meritocratic perceptions and meritocratic preferences, suggesting that students who perceive society as more meritocratic also tend to endorse meritocratic ideals. The strongest association is between privilege perceptions and meritocratic preferences, indicating that students who recognize the role of privilege in society are also more likely to endorse meritocratic ideals. However, other correlations suggest relationships that are less straightforward, as for instance the positive correlation between meritocratic preferences and privilege acceptance, which may indicate that despite perceiving meritocracy, students may still accept the legitimacy of privilege advantages. Something similar occurs with positive association between meritocratic and privilege preferences, which may suggest that students who endorse meritocratic ideals may also recognize the legitimacy of privilege advantages. Although these last correlations are modest in size, they highlight the complexity of students' beliefs about meritocracy and privilege, suggesting that these dimensions are not simply opposites but rather can coexist in students minds without necessarily undermining one another.    


## Stability of meritocracy and privilege beliefs 

This section objective is twofold. On the one hand, we aim at testing the validity of the measurement model through an assessment of it stability over time and cohorts. For this, we perform an assessment of model equivalence or _invariance_, which is a form of robustness check that allows us to determine whether the latent constructs are being measured in the same way across different groups or time points. Such evidence is crucial for ensuring that any observed differences in the latent constructs are not due to measurement artifacts but rather reflect valid differences[@putnick_measurement_2016] . The test of the equivalence of parameters in a confirmatory factor analysis follows a hierarchical approach, starting with configural invariance (same factor structure across groups or time points), then weak invariance (equal factor loadings), strong invariance (equal intercepts), and finally strict invariance (equal residual variances). Each level of invariance provides increasingly stringent evidence that the measurement model is stable across groups or time points. In this study, we first test for longitudinal invariance across the two waves of data collection, and then we test for cohort invariance between primary and secondary education students.

### Longitudinal invariance

```{r longitudinal-data, include=FALSE}
#| results: asis
#| echo: false
#| warning: false
#| output: false

# Crear base en formato wide con dummy de cohorte (0 = primaria, 1 = secundaria)
db_invariance <- db_long %>%
  group_by(id_estudiante) %>%
  mutate(
    cohort_level = first(cohort_level),
    cohort_dummy = case_when(
      cohort_level == "Primary" ~ 0,
      cohort_level == "Secondary" ~ 1,
      TRUE ~ NA_real_
    )
  ) %>%
  ungroup() %>%
  select(id_estudiante, cohort_dummy, ola, starts_with(c("perc", "pref"))) %>%
  pivot_wider(
    id_cols = c(id_estudiante, cohort_dummy),
    names_from = ola,
    names_glue = "{.value}{ola}",
    values_from = c(
      perc_effort, perc_talent,
      perc_rich_parents, perc_contact,
      pref_effort, pref_talent,
      pref_rich_parents, pref_contact
    )
  ) %>%
  na.omit() %>%
  rename_with(~ str_replace_all(., "_", ""))

# Reescalar variables y convertir a ordinales
db_invariance <- db_invariance %>% 
  mutate(
    across(
      .cols = -c(idestudiante, cohortdummy),
      .fns = ~ case_when(. == 1 ~ 0,
                         . == 2 ~ 1,
                         . == 3 ~ 2,
                         . == 4 ~ 3)
    )
  ) %>%
  mutate(
    across(
      .cols = -c(idestudiante),
      .fns = ~ ordered(.)
    )
  )
```

```{r, include=FALSE}
#| label: conf-model
#| tbl-cap: "Longitudinal Invariance Test"
#| results: asis
#| echo: false
#| warning: false

# First, define the configural model, using the repeated measures factors and
# indicators
baseline_model_smt <- ('

###########################################
# Definición de factores (1 marcador por factor)
###########################################

percmerit1  =~ 1*perceffort1     + perctalent1
percmerit2  =~ 1*perceffort2     + perctalent2

percnmerit1 =~ 1*percrichparents1 + perccontact1
percnmerit2 =~ 1*percrichparents2 + perccontact2

prefmerit1  =~ 1*prefeffort1     + preftalent1
prefmerit2  =~ 1*prefeffort2     + preftalent2

prefnmerit1 =~ 1*prefrichparents1 + prefcontact1
prefnmerit2 =~ 1*prefrichparents2 + prefcontact2


###########################################
# Covarianzas entre errores (mismo ítem, distintas olas)
###########################################

perceffort1 ~~ perceffort2
perctalent1 ~~ perctalent2

percrichparents1 ~~ percrichparents2
perccontact1     ~~ perccontact2

prefeffort1 ~~ prefeffort2
preftalent1 ~~ preftalent2

prefrichparents1 ~~ prefrichparents2
prefcontact1     ~~ prefcontact2
')



# Model Estimation
fit_baseline <- run_cfa(baseline_model_smt, data = db_invariance,
                    ordered = c("perceffort1", "perctalent1", "perceffort2", "perctalent2",
                                "percrichparents1", "perccontact1", "percrichparents2", "perccontact2",
                                "prefeffort1", "preftalent1", "prefeffort2", "preftalent2",
                                "prefrichparents1", "prefcontact1", "prefrichparents2", "prefcontact2"),
                    parameterization = "theta",
                    estimator = "WLSMV")
```

```{r, include=FALSE}
#| label: Weak-Model
#| tbl-cap: "Longitudinal Invariance Weak Model"
#| results: asis
#| echo: false
#| warning: false

loadinginv_model_smt <- ('

# Igualación de cargas (invarianza débil)

percmerit1  =~ 1*perceffort1 + pe_loading*perctalent1
percmerit2  =~ 1*perceffort2 + pe_loading*perctalent2

percnmerit1 =~ 1*percrichparents1 + prc_loading*perccontact1
percnmerit2 =~ 1*percrichparents2 + prc_loading*perccontact2

prefmerit1  =~ 1*prefeffort1 + pfe_loading*preftalent1
prefmerit2  =~ 1*prefeffort2 + pfe_loading*preftalent2

prefnmerit1 =~ 1*prefrichparents1 + pfrc_loading*prefcontact1
prefnmerit2 =~ 1*prefrichparents2 + pfrc_loading*prefcontact2


# Varianzas y covarianzas latentes

percmerit1  ~~ percmerit1 + percmerit2
percmerit2  ~~ percmerit2
percnmerit1 ~~ percnmerit1 + percnmerit2
percnmerit2 ~~ percnmerit2
prefmerit1  ~~ prefmerit1 + prefmerit2
prefmerit2  ~~ prefmerit2
prefnmerit1 ~~ prefnmerit1 + prefnmerit2
prefnmerit2 ~~ prefnmerit2


# Umbrales (uno igualado por ítem)

# perceffort1 | pe1t1*t1 + pe1t2*t2
# perceffort2 | pe2t1*t1 + pe2t2*t2
# perctalent1 | pt1t1*t1 + pt1t2*t2
# perctalent2 | pt2t1*t1 + pt2t2*t2
# percrichparents1 | prp1t1*t1 + prp1t2*t2
# percrichparents2 | prp2t1*t1 + prp2t2*t2
# perccontact1 | pc1t1*t1 + pc1t2*t2
# perccontact2 | pc2t1*t1 + pc2t2*t2
# prefeffort1 | pre1t1*t1 + pre1t2*t2
# prefeffort2 | pre2t1*t1 + pre2t2*t2
# preftalent1 | prt1t1*t1 + prt1t2*t2
# preftalent2 | prt2t1*t1 + prt2t2*t2
# prefrichparents1 | pfrp1t1*t1 + pfrp1t2*t2
# prefrichparents2 | pfrp2t1*t1 + pfrp2t2*t2
# prefcontact1 | pfc1t1*t1 + pfc1t2*t2
# prefcontact2 | pfc2t1*t1 + pfc2t2*t2
 

# Covarianzas entre errores

perceffort1 ~~ perceffort2
perctalent1 ~~ perctalent2
percrichparents1 ~~ percrichparents2
perccontact1 ~~ perccontact2
prefeffort1 ~~ prefeffort2
preftalent1 ~~ preftalent2
prefrichparents1 ~~ prefrichparents2
prefcontact1 ~~ prefcontact2
')


fit_loadinginv <- run_cfa(loadinginv_model_smt, data = db_invariance,
                      ordered = c("perceffort1", "perctalent1", "perceffort2", "perctalent2",
                                  "percrichparents1", "perccontact1", "percrichparents2", "perccontact2",
                                  "prefeffort1", "preftalent1", "prefeffort2", "preftalent2",
                                  "prefrichparents1", "prefcontact1", "prefrichparents2", "prefcontact2"),
                      parameterization = "theta",
                      estimator = "WLSMV")
```

```{r, include=FALSE}
#| label: strong-model
#| tbl-cap: "Longitudinal Invariance Strong Model"
#| results: asis
#| echo: false
#| warning: false

thresholdinv_model_smt <- ('

# Igualación de cargas (como en el modelo débil)

percmerit1  =~ 1*perceffort1 + pe_loading*perctalent1
percmerit2  =~ 1*perceffort2 + pe_loading*perctalent2

percnmerit1 =~ 1*percrichparents1 + prc_loading*perccontact1
percnmerit2 =~ 1*percrichparents2 + prc_loading*perccontact2

prefmerit1  =~ 1*prefeffort1 + pfe_loading*preftalent1
prefmerit2  =~ 1*prefeffort2 + pfe_loading*preftalent2

prefnmerit1 =~ 1*prefrichparents1 + pfrc_loading*prefcontact1
prefnmerit2 =~ 1*prefrichparents2 + pfrc_loading*prefcontact2


# Varianzas y covarianzas latentes

percmerit1  ~~ percmerit1 + percmerit2
percmerit2  ~~ percmerit2
percnmerit1 ~~ percnmerit1 + percnmerit2
percnmerit2 ~~ percnmerit2
prefmerit1  ~~ prefmerit1 + prefmerit2
prefmerit2  ~~ prefmerit2
prefnmerit1 ~~ prefnmerit1 + prefnmerit2
prefnmerit2 ~~ prefnmerit2


# Thresholds iguales entre olas (invarianza fuerte)

perceffort1     | pe_t1*t1 + pe_t2*t2
perceffort2     | pe_t1*t1 + pe_t2*t2
perctalent1     | pt_t1*t1 + pt_t2*t2
perctalent2     | pt_t1*t1 + pt_t2*t2
percrichparents1| prp_t1*t1 + prp_t2*t2
percrichparents2| prp_t1*t1 + prp_t2*t2
perccontact1    | pc_t1*t1 + pc_t2*t2
perccontact2    | pc_t1*t1 + pc_t2*t2
prefeffort1     | pre_t1*t1 + pre_t2*t2
prefeffort2     | pre_t1*t1 + pre_t2*t2
preftalent1     | prt_t1*t1 + prt_t2*t2
preftalent2     | prt_t1*t1 + prt_t2*t2
prefrichparents1| pfrp_t1*t1 + pfrp_t2*t2
prefrichparents2| pfrp_t1*t1 + pfrp_t2*t2
prefcontact1    | pfc_t1*t1 + pfc_t2*t2
prefcontact2    | pfc_t1*t1 + pfc_t2*t2


# Interceptos fijos

perceffort1 + perctalent1 ~ 0*1
perceffort2 + perctalent2 ~ 0*1
percrichparents1 + perccontact1 ~ 0*1
percrichparents2 + perccontact2 ~ 0*1
prefeffort1 + preftalent1 ~ 0*1
prefeffort2 + preftalent2 ~ 0*1
prefrichparents1 + prefcontact1 ~ 0*1
prefrichparents2 + prefcontact2 ~ 0*1


# Varianzas únicas

perceffort1 ~~ 1*perceffort1
perctalent1 ~~ 1*perctalent1
percrichparents1 ~~ 1*percrichparents1
perccontact1 ~~ 1*perccontact1
prefeffort1 ~~ 1*prefeffort1
preftalent1 ~~ 1*preftalent1
prefrichparents1 ~~ 1*prefrichparents1
prefcontact1 ~~ 1*prefcontact1

perceffort2 ~~ NA*perceffort2
perctalent2 ~~ NA*perctalent2
percrichparents2 ~~ NA*percrichparents2
perccontact2 ~~ NA*perccontact2
prefeffort2 ~~ NA*prefeffort2
preftalent2 ~~ NA*preftalent2
prefrichparents2 ~~ NA*prefrichparents2
prefcontact2 ~~ NA*prefcontact2


# Covarianzas entre errores longitudinales

perceffort1 ~~ perceffort2
perctalent1 ~~ perctalent2
percrichparents1 ~~ percrichparents2
perccontact1 ~~ perccontact2
prefeffort1 ~~ prefeffort2
preftalent1 ~~ preftalent2
prefrichparents1 ~~ prefrichparents2
prefcontact1 ~~ prefcontact2
')

fit_thresholdinv <- run_cfa(thresholdinv_model_smt, data = db_invariance,
                        ordered = c("perceffort1", "perctalent1", "perceffort2", "perctalent2",
                                    "percrichparents1", "perccontact1", "percrichparents2", "perccontact2",
                                    "prefeffort1", "preftalent1", "prefeffort2", "preftalent2",
                                    "prefrichparents1", "prefcontact1", "prefrichparents2", "prefcontact2"),
                        parameterization = "theta",
                        estimator = "WLSMV")
```

```{r, include=FALSE}
#| label: strict-model
#| tbl-cap: "Longitudinal Invariance Strict Model"
#| results: asis
#| echo: false
#| warning: false

uniquenessinv_model_smt <- ('

# Cargas factoriales (invarianza débil)

percmerit1  =~ 1*perceffort1 + pe_loading*perctalent1
percmerit2  =~ 1*perceffort2 + pe_loading*perctalent2

percnmerit1 =~ 1*percrichparents1 + prc_loading*perccontact1
percnmerit2 =~ 1*percrichparents2 + prc_loading*perccontact2

prefmerit1  =~ 1*prefeffort1 + pfe_loading*preftalent1
prefmerit2  =~ 1*prefeffort2 + pfe_loading*preftalent2

prefnmerit1 =~ 1*prefrichparents1 + pfrc_loading*prefcontact1
prefnmerit2 =~ 1*prefrichparents2 + pfrc_loading*prefcontact2


# Varianzas y covarianzas de factores latentes

percmerit1  ~~ percmerit1 + percmerit2
percmerit2  ~~ percmerit2
percnmerit1 ~~ percnmerit1 + percnmerit2
percnmerit2 ~~ percnmerit2
prefmerit1  ~~ prefmerit1 + prefmerit2
prefmerit2  ~~ prefmerit2
prefnmerit1 ~~ prefnmerit1 + prefnmerit2
prefnmerit2 ~~ prefnmerit2


# Medias latentes

percmerit1  ~ 0*1
percmerit2  ~ 1
percnmerit1 ~ 0*1
percnmerit2 ~ 1
prefmerit1  ~ 0*1
prefmerit2  ~ 1
prefnmerit1 ~ 0*1
prefnmerit2 ~ 1


# Umbrales (invarianza fuerte)

perceffort1     | pe_t1*t1 + pe_t2*t2
perceffort2     | pe_t1*t1 + pe_t2*t2
perctalent1     | pt_t1*t1 + pt_t2*t2
perctalent2     | pt_t1*t1 + pt_t2*t2
percrichparents1| prp_t1*t1 + prp_t2*t2
percrichparents2| prp_t1*t1 + prp_t2*t2
perccontact1    | pc_t1*t1 + pc_t2*t2
perccontact2    | pc_t1*t1 + pc_t2*t2
prefeffort1     | pre_t1*t1 + pre_t2*t2
prefeffort2     | pre_t1*t1 + pre_t2*t2
preftalent1     | prt_t1*t1 + prt_t2*t2
preftalent2     | prt_t1*t1 + prt_t2*t2
prefrichparents1| pfrp_t1*t1 + pfrp_t2*t2
prefrichparents2| pfrp_t1*t1 + pfrp_t2*t2
prefcontact1    | pfc_t1*t1 + pfc_t2*t2
prefcontact2    | pfc_t1*t1 + pfc_t2*t2


# Interceptos fijos

perceffort1 + perctalent1 ~ 0*1
perceffort2 + perctalent2 ~ 0*1
percrichparents1 + perccontact1 ~ 0*1
percrichparents2 + perccontact2 ~ 0*1
prefeffort1 + preftalent1 ~ 0*1
prefeffort2 + preftalent2 ~ 0*1
prefrichparents1 + prefcontact1 ~ 0*1
prefrichparents2 + prefcontact2 ~ 0*1


# Varianzas únicas (invarianza estricta)
# Fijadas a 1 en todas las olas

perceffort1 ~~ 1*perceffort1
perceffort2 ~~ 1*perceffort2
perctalent1 ~~ 1*perctalent1
perctalent2 ~~ 1*perctalent2

percrichparents1 ~~ 1*percrichparents1
percrichparents2 ~~ 1*percrichparents2
perccontact1 ~~ 1*perccontact1
perccontact2 ~~ 1*perccontact2

prefeffort1 ~~ 1*prefeffort1
prefeffort2 ~~ 1*prefeffort2
preftalent1 ~~ 1*preftalent1
preftalent2 ~~ 1*preftalent2

prefrichparents1 ~~ 1*prefrichparents1
prefrichparents2 ~~ 1*prefrichparents2
prefcontact1 ~~ 1*prefcontact1
prefcontact2 ~~ 1*prefcontact2


# Covarianzas entre errores (mismo ítem, diferente ola)

perceffort1 ~~ perceffort2
perctalent1 ~~ perctalent2
percrichparents1 ~~ percrichparents2
perccontact1 ~~ perccontact2
prefeffort1 ~~ prefeffort2
preftalent1 ~~ preftalent2
prefrichparents1 ~~ prefrichparents2
prefcontact1 ~~ prefcontact2
')

fit_uniquenessinv <- run_cfa(uniquenessinv_model_smt, data = db_invariance,
                         ordered = c("perceffort1", "perctalent1", "perceffort2", "perctalent2",
                                     "percrichparents1", "perccontact1", "percrichparents2", "perccontact2",
                                     "prefeffort1", "preftalent1", "prefeffort2", "preftalent2",
                                     "prefrichparents1", "prefcontact1", "prefrichparents2", "prefcontact2"),
                         parameterization = "theta",
                         estimator = "WLSMV")
```

@tbl-longinv reports the results of the longitudinal invariance tests. The logic of longitudinal invariance is to apply senquential constraints to the model parameters across time, in order to check to what extent the parameters are invariant across time. This is basically takign the two measurement points as two comparing groups, but in this case, the groups are the same individuals measured at two different time points, which is incorporated in the model by correlating the unique item variances. The table shows the fit indices for the configural, weak (metric), strong (scalar), and strict invariance models, along with the chi-square difference tests comparing each nested model. Overall, the scale achieved the strongest level of invariance (strict invariance) across the two waves. The strict model showed good fit, $\chi^2$(84) = 130.17, CFI = 0.991, RMSEA = 0.031 (90% CI: 0.020–0.040). Moreover, compared with the strong (scalar) model, the deterioration in fit was negligible: $\Delta \chi^2$(4) = 1.635, $\Delta$CFI = 0.000, and $\Delta$RMSEA = -0.002. These differences fall well within commonly used cutoffs for invariance evaluation [@chen_sensitivity_2007], indicating that constraining residual variances in addition to loadings and thresholds/intercepts is tenable. Substantively, this implies that the item–factor relations are stable over time (metric invariance) and that, for a given level of the latent trait, respondents are expected to endorse the same response categories across waves (scalar invariance), supporting the comparability of scores and latent parameters within students over time.

```{r}
#| tbl-cap: "Longitudinal invariance results"
#| tbl-cap-location: top
#| tbl-pos: H
#| echo: false
#| warning: false
#| label: tbl-longinv

# Compare fit

an1 <- anova(fit_baseline, fit_loadinginv)
an2 <- anova(fit_loadinginv, fit_thresholdinv)
an3 <- anova(fit_thresholdinv, fit_uniquenessinv)

tab01 <- bind_rows(
  as_tibble(an1)[2,],
  as_tibble(an2)[2,],
  as_tibble(an3)[2,]
) %>%
  select("Chisq", "Df", chisq_diff = `Chisq diff`, df_diff = `Df diff`, pvalue = `Pr(>Chisq)`) %>%
  mutate(
    stars = stars.pval(pvalue),
    chisqt = paste0(round(Chisq, 2), " (", Df, ")"),
    decision = ifelse(pvalue > 0.05, "Accept", "Reject"),
    model = c("Weak", "Strong", "Strict")
  ) %>%
  bind_rows(
    tibble(Chisq = an1$Chisq[1], Df = an1$Df[1], chisq_diff = NA, df_diff = NA,
           pvalue = NA, stars = "", chisqt = paste0(round(an1$Chisq[1], 2), " (", an1$Df[1], ")"),
           decision = "Reference", model = "Configural")
  ) %>%
  select(model, chisqt, chisq_diff, df_diff, pvalue, stars, decision) %>%
  mutate(model = factor(model, levels = c("Configural", "Weak", "Strong", "Strict"))) %>%
  arrange(model)


fit.meas <- dplyr::bind_rows(lavaan::fitmeasures(fit_baseline, output ="matrix")[c("chisq","df","cfi","rmsea","rmsea.ci.lower","rmsea.ci.upper"),],
                             lavaan::fitmeasures(fit_loadinginv,  output ="matrix")[c("chisq","df","cfi","rmsea","rmsea.ci.lower","rmsea.ci.upper"),],
                             lavaan::fitmeasures(fit_thresholdinv, output ="matrix")[c("chisq","df","cfi","rmsea","rmsea.ci.lower","rmsea.ci.upper"),],
                             lavaan::fitmeasures(fit_uniquenessinv, output ="matrix")[c("chisq","df","cfi","rmsea","rmsea.ci.lower","rmsea.ci.upper"),])

fit.meas <- fit.meas %>%
  dplyr::mutate(diff.chi2 = chisq    - lag(chisq,default = dplyr::first(chisq)),
                diff.df   = df       - lag(df,   default = dplyr::first(df)),
                diff.cfi  = cfi      - lag(cfi,  default = dplyr::first(cfi)),
                diff.rmsea   = rmsea - lag(rmsea,default = dplyr::first(rmsea))) %>%
  round(3) %>%
  dplyr::mutate(rmsea.ci=paste0(rmsea," \n ", "(",rmsea.ci.lower,"-",rmsea.ci.upper,")"))

tab.inv <- dplyr::bind_cols(tab01,fit.meas) %>%
  dplyr::select(model,chisqt,cfi,rmsea.ci,diff.chi2,diff.df,diff.cfi,diff.rmsea,stars,decision) %>%
  dplyr::mutate(diff.chi2=paste0(diff.chi2," (",diff.df,") ",stars)) %>%
  dplyr::select(model,chisqt,cfi,rmsea.ci,diff.chi2,diff.cfi,diff.rmsea,decision)


col.nam <- if (is_html_output()) {
  c("Model", "&chi;<sup>2</sup> (df)", "CFI", "RMSEA (90% CI)",
    "&Delta;&chi;<sup>2</sup> (&Delta;df)", "&Delta;CFI", "&Delta;RMSEA", "Decision")
} else {
  c("Model", "$\\chi^2$ (df)", "CFI", "RMSEA (90\\% CI)",
    "$\\Delta \\chi^2$ ($\\Delta$df)", "$\\Delta$CFI", "$\\Delta$RMSEA", "Decision")
}

tab.inv %>%
  kableExtra::kable(format = table_format,
                    align = "c",
                    booktabs = T,
                    escape = F,
                    caption = NULL,
                    col.names = col.nam) %>%
  kableExtra::kable_styling(full_width = F,
                            latex_options = c("HOLD_position", "scale_down"),
                            bootstrap_options=c("striped", "bordered", "condensed"),
                           font_size = 9.5) %>%
  kableExtra::column_spec(column = 1, width = "2cm") %>%
  kableExtra::column_spec(column = 2, width = "2.1cm") %>%
  kableExtra::column_spec(column = 3, width = "1.3cm") %>%
  kableExtra::column_spec(column = 4, width = "2.2cm") %>%
  kableExtra::column_spec(column = 5, width = "2.4cm") %>%
  kableExtra::column_spec(column = 6, width = "1.3cm") %>%
  kableExtra::column_spec(column = 7, width = "1.4cm") %>%
  kableExtra::column_spec(column = 8, width = "2cm")
```
As an additional diagnostic, we used univariate score tests to identify potential sources of localized misfit at each step of the invariance hierarchy. The tests revealed no meaningful violations of the imposed equality constraints, suggesting that model fit was not driven by a small subset of parameters. Global score tests and the smallest p-values at each step are reported in the Supplementary Material. Based on these results, we conclude that the measurement model is stable over time, and that the latent constructs are being measured in the same way across the two waves of data collection.

## Cohorts comparability


```{r}
#| label: trad-cohort-invariance
#| include: false

# Traditional multigroup invariance with native lavaan (wave 1, primary vs secondary)
# For ordinal indicators: configural -> loadings -> loadings + thresholds

db_trad <- db_long %>%
  filter(ola == 1) %>%
  select(id_estudiante, starts_with(c("perc", "pref")), cohort_level) %>%
  drop_na()

model_cohort <- '
perc_merit =~ perc_effort + perc_talent
perc_nmerit =~ perc_rich_parents + perc_contact
pref_merit =~ pref_effort + pref_talent
pref_nmerit =~ pref_rich_parents + pref_contact
'

ord_items <- c("perc_effort", "perc_talent", "perc_rich_parents", "perc_contact",
               "pref_effort", "pref_talent", "pref_rich_parents", "pref_contact")

trad_configural <- cfa(model_cohort, data = db_trad, group = "cohort_level",
                       ordered = ord_items, estimator = "WLSMV")

trad_loadings <- cfa(model_cohort, data = db_trad, group = "cohort_level",
                     ordered = ord_items, estimator = "WLSMV",
                     group.equal = "loadings")

trad_thresholds <- cfa(model_cohort, data = db_trad, group = "cohort_level",
                       ordered = ord_items, estimator = "WLSMV",
                       group.equal = c("loadings", "thresholds"))

# chi-square difference tests and fit indices
lavTestLRT(trad_configural, trad_loadings, trad_thresholds)

sapply(list(configural = trad_configural, loadings = trad_loadings, thresholds = trad_thresholds),
       fitMeasures, fit.measures = c("chisq.scaled", "df.scaled", "cfi.scaled", "rmsea.scaled"))

#                 Df AIC BIC  Chisq Chisq diff    RMSEA Df diff Pr(>Chisq)    
# trad_configural 28         30.368                                           
# trad_loadings   32         48.890     20.769 0.092642       4  0.0003518 ***
# trad_thresholds 40         68.130     35.890 0.057633       8  1.839e-05 ***

# No invariance

# --- Sources of non-invariance ---------------------------------------------
# 1) Score tests on the loadings-constrained model: one test per equality constraint
score_loadings <- lavTestScore(trad_loadings)
score_loadings$test                 # joint test of all constraints

pt <- parTable(trad_loadings)       # map constraint labels (.pN.) to parameters
score_loadings$uni %>%
  left_join(pt %>% filter(group == 1) %>% select(plabel, lhs.par = lhs, rhs.par = rhs),
            by = c("lhs" = "plabel")) %>%
  mutate(p.holm = p.adjust(p.value, "holm")) %>%
  arrange(p.value)

# 2) Group-specific loadings in the configural model (where do they differ?)
parameterEstimates(trad_configural, standardized = TRUE) %>%
  filter(op == "=~") %>%
  select(lhs, rhs, group, est, std.all)

# 3) Partial invariance: free the offending loading (pref_talent), keep the rest equal
trad_partial_load <- cfa(model_cohort, data = db_trad, group = "cohort_level",
                         ordered = ord_items, estimator = "WLSMV",
                         group.equal = "loadings",
                         group.partial = "pref_merit =~ pref_talent")

trad_partial_thres <- cfa(model_cohort, data = db_trad, group = "cohort_level",
                          ordered = ord_items, estimator = "WLSMV",
                          group.equal = c("loadings", "thresholds"),
                          group.partial = "pref_merit =~ pref_talent")

lavTestLRT(trad_configural, trad_partial_load, trad_partial_thres)

sapply(list(configural = trad_configural, partial_loadings = trad_partial_load,
            partial_thresholds = trad_partial_thres),
       fitMeasures, fit.measures = c("chisq.scaled", "df.scaled", "cfi.scaled", "rmsea.scaled"))

lavTestScore(trad_partial_thres)$test   # joint test of the remaining constraints

# 4) Which thresholds drive the misfit? Univariate score tests by item
score_thres <- lavTestScore(trad_partial_thres)
lab_thres <- parTable(trad_partial_thres) %>%
  filter(group == 1) %>%
  transmute(plabel, item = lhs, threshold = rhs)

score_thres$uni %>%
  left_join(lab_thres, by = c("lhs" = "plabel")) %>%
  mutate(p.holm = p.adjust(p.value, "holm")) %>%
  arrange(p.value) %>%
  select(item, threshold, X2, p.value, p.holm)
```

@tbl-cohortinv reports the results of the cross-cohort (multigroup) invariance tests comparing primary and secondary students. As in the longitudinal case, the sequence starts from a configural model, which imposes the same four-factor structure in both cohorts but lets all parameters vary freely, and then adds equality constraints on factor loadings (weak invariance) and on loadings and item thresholds (strong invariance). Since the indicators are ordinal, thresholds play the role that intercepts play for continuous items. The configural model shows a good fit, $\chi^2$(28) = 47.65, CFI = 0.991, RMSEA = 0.041, indicating that the same four-factor structure holds in both cohorts. However, the scale did not achieve weak invariance: constraining the factor loadings to be equal significantly worsened fit, $\Delta \chi^2$(4) = 20.77, $p$ < .001. This means that the relationship between the latent factors and their indicators differs between primary and secondary students, which could be due to developmental differences or differences in how the items are interpreted across cohorts. An additional analysis following the alternative sequence proposed by Svetina et al. [-@svetina_multiplegroup_2020], which tests the equivalence of thresholds before loadings, led to the same conclusion: invariance between primary and secondary students was not achieved either (see supplementary materials).

The changes in CFI ($\Delta$CFI = −0.009) and RMSEA ($\Delta$RMSEA = 0.013) are close to, though still within, the conventional cutoffs [@chen_sensitivity_2007], so at the global level the evidence against loading invariance is not overwhelming. Adding the threshold constraints worsened fit further, $\Delta \chi^2$(8) = 35.89, $p$ < .001, with $\Delta$CFI = −0.011, which exceeds the recommended cutoff.

```{r}
#| tbl-cap: "Cohort (multigroup) invariance results"
#| tbl-cap-location: top
#| tbl-pos: H
#| echo: false
#| warning: false
#| label: tbl-cohortinv

# Models and the model each one is compared against (NA = reference)
fit_coh <- list(
  "Configural"                     = trad_configural,
  "Loadings"                       = trad_loadings,
  "Loadings + Thresholds"          = trad_thresholds,
  "Loadings (partial)"             = trad_partial_load,
  "Loadings (partial) + Thresholds" = trad_partial_thres
)
ref_coh <- c(NA, 1, 2, 1, 4)

fi_coh <- c("chisq.scaled", "df.scaled", "cfi.scaled", "rmsea.scaled",
            "rmsea.ci.lower.scaled", "rmsea.ci.upper.scaled")

fm_coh <- t(sapply(fit_coh, fitMeasures, fit.measures = fi_coh)) %>%
  as.data.frame() %>%
  rename_with(~ str_remove(., ".scaled"))

# scaled chi-square difference tests against the reference model of each row
lrt_coh <- purrr::map_dfr(seq_along(fit_coh), function(i) {
  if (is.na(ref_coh[i])) return(tibble(dchi = NA_real_, ddf = NA_real_, pvalue = NA_real_))
  a <- suppressMessages(suppressWarnings(
    lavTestLRT(fit_coh[[ref_coh[i]]], fit_coh[[i]])[2, ]
  ))
  tibble(dchi = a$`Chisq diff`, ddf = a$`Df diff`, pvalue = a$`Pr(>Chisq)`)
})

tab_coh <- bind_cols(model = names(fit_coh), fm_coh, lrt_coh) %>%
  mutate(
    diff.cfi   = cfi - cfi[ref_coh],
    diff.rmsea = rmsea - rmsea[ref_coh],
    chisqt     = paste0(round(chisq, 2), " (", df, ")"),
    rmsea.ci   = paste0(round(rmsea, 3), " (", round(rmsea.ci.lower, 3), "-", round(rmsea.ci.upper, 3), ")"),
    diff.chi2  = ifelse(is.na(dchi), "",
                        paste0(round(dchi, 2), " (", ddf, ") ", stars.pval(pvalue))),
    decision   = case_when(is.na(pvalue) ~ "Reference",
                           pvalue > 0.05 ~ "Accept",
                           TRUE ~ "Reject"),
    cfi        = round(cfi, 3),
    diff.cfi   = ifelse(is.na(dchi), "", format(round(diff.cfi, 3), nsmall = 3)),
    diff.rmsea = ifelse(is.na(dchi), "", format(round(diff.rmsea, 3), nsmall = 3))
  ) %>%
  select(model, chisqt, cfi, rmsea.ci, diff.chi2, diff.cfi, diff.rmsea, decision)

col.nam <- if (is_html_output()) {
  c("Model", "&chi;<sup>2</sup> (df)", "CFI", "RMSEA (90% CI)",
    "&Delta;&chi;<sup>2</sup> (&Delta;df)", "&Delta;CFI", "&Delta;RMSEA", "Decision")
} else {
  c("Model", "$\\chi^2$ (df)", "CFI", "RMSEA (90\\% CI)",
    "$\\Delta \\chi^2$ ($\\Delta$df)", "$\\Delta$CFI", "$\\Delta$RMSEA", "Decision")
}

tab_coh_kbl <- tab_coh %>%
  kableExtra::kable(format = table_format,
                    align = "c",
                    booktabs = T,
                    escape = F,
                    caption = NULL,
                    col.names = col.nam) %>%
  kableExtra::kable_styling(full_width = F,
                            latex_options = c("HOLD_position", "scale_down"),
                            bootstrap_options = c("striped", "bordered", "condensed"),
                            font_size = 9.5) %>%
  kableExtra::pack_rows("Full invariance sequence", 1, 3) %>%
  kableExtra::pack_rows("Partial invariance (loading of preferences for talent freed)", 4, 5) %>%
  kableExtra::column_spec(column = 1, width = "3.6cm") %>%
  kableExtra::column_spec(column = 2, width = "2.1cm") %>%
  kableExtra::column_spec(column = 3, width = "1.3cm") %>%
  kableExtra::column_spec(column = 4, width = "2.2cm") %>%
  kableExtra::column_spec(column = 5, width = "2.4cm") %>%
  kableExtra::column_spec(column = 6, width = "1.3cm") %>%
  kableExtra::column_spec(column = 7, width = "1.4cm") %>%
  kableExtra::column_spec(column = 8, width = "2cm")

tab_note <- "Scaled WLSMV statistics. Each model is compared with the preceding one in its block (partial loadings: vs. configural)."

# note below the table, outside the cells, in a smaller font than the table (9.5pt)
if (is_html_output()) {
  knitr::asis_output(paste0(
    tab_coh_kbl,
    "<div style='font-size: 0.75em; text-align: left; margin-top: -0.9em;'><em>Note.</em> ",
    tab_note, "</div>"
  ))
} else {
  knitr::asis_output(paste0(
    tab_coh_kbl,
    "\n\\vspace{-1.2em}\n\\begingroup\\fontsize{8}{9.5}\\selectfont\\raggedright ",
    "\\textit{Note.} ", tab_note, "\\par\\endgroup\n"
  ))
}
```

To identify the source of non-invariance, we inspected univariate score tests for each equality constraint on the loadings. Unlike in the longitudinal case, the misfit was concentrated in a single parameter: the loading of the talent preference item on the meritocratic preferences factor ($\chi^2$(1) = 16.87, Holm-adjusted $p$ < .001). The remaining loadings were fully consistent with invariance (all $p$ > .36). In line with this, freeing that single loading is enough to restore weak invariance, with a fit indistinguishable from the configural model, $\Delta \chi^2$(3) = 1.82, $p$ = .61, CFI = 0.992, RMSEA = 0.037.

The cohort-specific loadings show what this means substantively, as revealed when observing the factor models separately for each cohort (see supplementary materialss). In primary school the meritocratic preferences factor is defined mainly by the preference for talent as a criterion of reward (standardized loading = 0.87) and much less by the preference for effort (0.45). In secondary school the pattern reverses: effort becomes the core of the factor (0.83) whereas talent is only weakly related to it (0.37). Thus, younger students seem to treat effort and talent as two sides of the same meritocratic ideal, whereas older students distinguish them, with effort remaining the central meritocratic criterion and talent becoming a more separate consideration, plausibly as inherited or unearned advantages become more salient with age. This means that the meritocratic preferences factor does not have exactly the same meaning in both cohorts, and comparisons of its levels across cohorts should be interpreted with caution. In contrast, the loadings for the perceptions of meritocracy and for both privilege factors are equivalent across cohorts.

Even after freeing this loading, adding the threshold constraints significantly worsened fit ($\Delta \chi^2$(8) = 42.72, $p$ < .001; $\Delta$CFI = −0.014), so strong invariance was not achieved. In this case, the source is different: univariate score tests point to the thresholds of the perceived effort item, for which all three constraints are violated ($\chi^2$(1) between 11.6 and 11.9; Holm-adjusted $p$ between .015 and .017), whereas no threshold of any other item shows meaningful misfit (adjusted $p$ > .99). Substantively, students in the two cohorts use the response categories of this item differently for a given level of perceived meritocracy, which is consistent with the descriptive differences reported above (see @fig-likerplot2): secondary students are more likely to endorse the lowest response category when judging whether effort is rewarded in society, suggesting that they apply a more critical standard. Taken together, only partial weak invariance holds, and comparisons of latent means across cohorts, in particular for meritocratic perceptions, should be interpreted with caution.

