---
title: "Results: Latent Class Analysis"
date: today
format: html
code-fold: true
toc: true
---

## Processing

Load libraries

```{r}
pacman::p_load(tidyverse,
               sjmisc,
               here,
               lavaan,
               psych,
               corrplot,
               ggdist,
               sjlabelled,
               patchwork,
               RColorBrewer,
               poLCA,
               reshape,
               summarytools,
               plotly)

options(scipen=999)
rm(list = ls())
```

Load database

```{r}
load("../output/data/db_proc.RData")
```

Items dichotomization

```{r}
#| warning: false

vars_m <- c("perc_effort",
            "perc_talent",
            "perc_rich_parents",
            "perc_contact",
            "pref_effort",
            "pref_talent",
            "pref_rich_parents",
            "pref_contact")

labels1 <- c("Strongly desagree" = 1, 
             "Desagree" = 2, 
             "Agree" = 3, 
             "Strongly agree" = 4)
db <- db %>% 
  mutate_at(.vars = all_of(vars_m),.funs = ~ sjlabelled::set_labels(., labels = labels1))

db <- db %>% 
  mutate(across(
    all_of(vars_m),
    ~ if_else(. >= 3, 1, 0),
    .names = "{.col}_d"   # perc_effort -> perc_effort_d, etc.
  ))

db <- db %>% 
  mutate(
    across(.cols = ends_with("_d"),
           .fns = ~ as.factor(.))
  )

db <- db %>% 
  mutate_at(.vars = all_of(vars_m),
            .funs = ~ factor(., 
                             levels = 1:4, 
                             labels = c("Strongly desagree",
                                        "Desagree",                  
                                        "Agree",                     
                                        "Strongly agree")))


db <- db %>% 
  mutate(
    across(
      ends_with("_d"),
      .fns = ~ factor(., 
                       levels = 0:1, 
                       labels = c("Low", "High"))))
   
```

Items labels

```{r}
db$perc_effort <- sjlabelled::set_label(db$perc_effort, 
                                                label = "In Chile people are rewarded for their efforts")


db$perc_talent <- sjlabelled::set_label(db$perc_talent, 
                                        label = "In Chile people are rewarded for their intelligence and ability")

db$perc_rich_parents <- sjlabelled::set_label(db$perc_rich_parents, 
                                        label = "In Chile those with wealthy parents do much better in life")

db$perc_contact <- sjlabelled::set_label(db$perc_contact, 
                                        label = "In Chile those with good contacts do much better in life")


db$perc_effort_d <- sjlabelled::set_label(db$perc_effort_d, 
                                          label = "In Chile people are rewarded for their efforts")


db$perc_talent_d <- sjlabelled::set_label(db$perc_talent_d, 
                                          label = "In Chile people are rewarded for their intelligence and ability")

db$perc_rich_parents_d <- sjlabelled::set_label(db$perc_rich_parents_d, 
                                                label = "In Chile those with wealthy parents do much better in life")

db$perc_contact_d <- sjlabelled::set_label(db$perc_contact_d, 
                                           label = "In Chile those with good contacts do much better in life")

db$pref_effort <- sjlabelled::set_label(db$pref_effort, 
                                        label = "Those who work harder should reap greater rewards than those who work less hard")


db$pref_talent <- sjlabelled::set_label(db$pref_talent, 
                                        label = "Those with more talent should reap greater rewards than those with less talent")

db$pref_rich_parents <- sjlabelled::set_label(db$pref_rich_parents, 
                                              label = "It is good that those who have rich parents do better in life")

db$pref_contact <- sjlabelled::set_label(db$pref_contact, 
                                         label = "It is good that those who have good contacts do better in life")


db$pref_effort_d <- sjlabelled::set_label(db$pref_effort_d, 
                                          label = "Those who work harder should reap greater rewards than those who work less hard")


db$pref_talent_d <- sjlabelled::set_label(db$pref_talent_d, 
                                          label = "Those with more talent should reap greater rewards than those with less talent")

db$pref_rich_parents_d <- sjlabelled::set_label(db$pref_rich_parents_d, 
                                                label = "It is good that those who have rich parents do better in life")

db$pref_contact_d <- sjlabelled::set_label(db$pref_contact_d, 
                                           label = "It is good that those who have good contacts do better in life")
```



## Latent class analysis (dichotomized items)

### Estimation

```{r}
#| include: false
# --- helpers -------------------------------------------------------------

get_K <- function(fit) {
  # robust: number of classes = columns of posterior
  if (!is.null(fit$posterior)) return(as.integer(ncol(fit$posterior)))
  # fallback: rows of probs for first item
  if (!is.null(fit$probs) && length(fit$probs) > 0) return(as.integer(nrow(fit$probs[[1]])))
  stop("Cannot determine K from poLCA object (no posterior/probs).")
}

get_npar <- function(fit) {
  if (!is.null(fit$npar)) return(as.numeric(fit$npar))
  # fallback if npar missing (rare): compute from probs + class shares
  K <- get_K(fit)
  Rj <- vapply(fit$probs, function(m) if (is.matrix(m)) ncol(m) else NA_integer_, integer(1))
  if (anyNA(Rj)) stop("Cannot compute npar: missing category counts in fit$probs.")
  as.numeric(sum(K * (Rj - 1)) + (K - 1))
}

get_n_patterns_obs <- function(fit) {
  # poLCA usually stores predicted cell counts for observed patterns as a vector
  if (!is.null(fit$predcell)) return(as.integer(length(fit$predcell)))
  # fallback: sometimes observed pattern table exists
  if (!is.null(fit$observed) && is.matrix(fit$observed)) return(as.integer(nrow(fit$observed)))
  NA_integer_
}

n_patterns_from_data <- function(data_indicators) {
  # data_indicators = db_lca with only indicators (your a1..a8), already NA-omitted
  as.integer(nrow(dplyr::distinct(data_indicators)))
}


entropy_stats <- function(fit) {
  P <- fit$posterior
  P <- pmax(P, 1e-12)
  N <- nrow(P); K <- ncol(P)
  
  # relative entropy (0..1): 1 = best separation
  rel_entropy <- 1 - (-sum(P * log(P)) / (N * log(K)))
  
  class_hat <- max.col(P)
  max_p <- apply(P, 1, max)
  
  avg_diag <- vapply(1:K, function(k) {
    idx <- which(class_hat == k)
    if (length(idx) == 0) NA_real_ else mean(P[idx, k])
  }, numeric(1))
  
  tibble(
    entropy = rel_entropy,
    mean_max_p = mean(max_p),
    p10_max_p = unname(quantile(max_p, 0.10)),
    avg_diag_min = suppressWarnings(min(avg_diag, na.rm = TRUE)),
    avg_diag_mean = mean(avg_diag, na.rm = TRUE)
  )
}

min_class_share <- function(fit) {
  Ppost <- fit$posterior
  N <- nrow(Ppost); K <- ncol(Ppost)
  class_hat <- max.col(Ppost)
  
  tibble(
    min_class_prior = min(fit$P),
    min_class_map   = min(tabulate(class_hat, nbins = K)) / N
  )
}

```


```{r}
#| include: false
# 3.3 LCA: dichotomized items ----

df <- db %>% 
  dplyr::select(ends_with("_d")) %>% 
  na.omit()

a1 <- df$perc_effort_d
a2 <- df$perc_talent_d
a3 <- df$perc_rich_parents_d
a4 <- df$perc_contact_d
a5 <- df$pref_effort_d
a6 <- df$pref_talent_d 
a7 <- df$pref_rich_parents_d
a8 <- df$pref_contact_d

db_lca <- data.frame(a1, a2, a3, a4, a5, a6, a7, a8) %>% as_tibble()

db_lca <- db_lca %>%
  mutate_at(.vars = 1:8, ~as.factor(.))

f <-cbind(a1, a2, a3, a4, a5, a6, a7, a8)~1

fit_lca <- function(K, data, formula, nrep = 50, maxiter = 3000, seed = 123) {
  set.seed(seed)
  poLCA(formula, data = data, nclass = K,
        nrep = nrep, maxiter = maxiter,
        verbose = FALSE, graphs = FALSE)
}

Ks <- 1:5
#fits <- lapply(Ks, fit_lca, data = db_lca, formula = f, nrep = 30)
#save(fits, file = here("output/fits_models_lca.RData"))

load(file = here("output/fits_models_lca.RData"))

fit_tbl <- data.frame(
  K    = Ks,
  logLik = sapply(fits, `[[`, "llik"),
  AIC  = sapply(fits, `[[`, "aic"),
  BIC  = sapply(fits, `[[`, "bic"),
  Gsq  = sapply(fits, `[[`, "Gsq"),
  Chisq = sapply(fits, `[[`, "Chisq")
)

fit_tbl
```


```{r}
#| label: tbl-fit
#| tbl-cap: "Fit comparision of latent class models"
#| tbl-cap-location: top

# ---- build table (no list-cols) -----------------------------------------

N_used <- nrow(db_lca)

m_patterns <- n_patterns_from_data(db_lca)  # scalar

# Robust extractors (avoid list-columns)
get_K <- function(fit) as.integer(ncol(fit$posterior))  # classes = cols of posterior
get_ll  <- function(fit) as.numeric(fit$llik)
get_aic <- function(fit) as.numeric(fit$aic)
get_bic <- function(fit) as.numeric(fit$bic)
get_npar <- function(fit) {
  if (!is.null(fit$npar)) return(as.numeric(fit$npar))
  # fallback: compute from probs
  K <- get_K(fit)
  Rj <- vapply(fit$probs, function(m) ncol(m), integer(1))
  as.numeric(sum(K * (Rj - 1)) + (K - 1))
}

# Table of fit indices
fit_tbl <- tibble(
  K     = map_int(fits, get_K),
  N     = nrow(db_lca),
  logLik= map_dbl(fits, get_ll),
  npar  = map_dbl(fits, get_npar),
  AIC   = map_dbl(fits, get_aic),
  BIC   = map_dbl(fits, get_bic)
) %>%
  arrange(K) %>%
  mutate(
    # Differences relative to preceding (K-1) model
    dAIC = AIC - lag(AIC),
    dBIC = BIC - lag(BIC),
    
    # Likelihood-ratio test (K vs K-1)
    # Test statistic: 2*(LL_K - LL_{K-1}) ~ Chi-square with df = npar_K - npar_{K-1}
    # Note: in LCA, models are not strictly nested in the regular sense;
    # this LRT is widely used as a heuristic, but interpret with caution.
    LRT = 2 * (logLik - lag(logLik)),
    df_LRT = npar - lag(npar),
    p_LRT = ifelse(!is.na(LRT) & df_LRT > 0, pchisq(LRT, df = df_LRT, lower.tail = FALSE), NA_real_)
  )


fit_tbl_report <- fit_tbl %>%
  transmute(
    K, N,
    logLik = round(logLik, 1),
    npar = npar,
    AIC = round(AIC, 1),
    dAIC = round(dAIC, 1),
    BIC = round(BIC, 1),
    dBIC = round(dBIC, 1),
    LRT = round(LRT, 1),
    df_LRT = df_LRT,
    p_LRT = signif(p_LRT, 3)
  )

fit_tbl_report %>% 
 kableExtra::kable(format = "html",
                    align = "c",
                    booktabs = T,
                    escape = F,
                    caption = NULL) %>%
  kableExtra::kable_styling(full_width = T,
                            latex_options = "hold_position",
                            bootstrap_options=c("striped", "bordered", "condensed"),
                            font_size = 23) %>%
  kableExtra::column_spec(c(1,8), width = "3.5cm") %>% 
  kableExtra::column_spec(2:7, width = "4cm") %>% 
  kableExtra::column_spec(4, width = "5cm")

```

```{r}
#| label: fig-visaic
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "Visual: AIC/BIC by K (lower is better)"
#| fig-cap-location: top


# Visual: AIC/BIC by K (lower is better)
fit_long <- fit_tbl %>%
  dplyr::select(K, AIC, BIC) %>%
  pivot_longer(-K, names_to = "criterion", values_to = "value")

ggplot(fit_long, aes(x = K, y = value, group = criterion)) +
  geom_line() +
  geom_point() +
  facet_wrap(~criterion, scales = "free_y") +
  labs(x = "Number of classes (K)", y = "Value")


```

```{r}
#| label: fig-dfvisaic
#| out-width: '80%'
#| fig-asp: 0.8
#| fig-cap: "ΔAIC and ΔBIC (negative = improvement over K-1)"
#| fig-cap-location: top
#| warning: false


# Visual: ΔAIC and ΔBIC (negative = improvement over K-1)
delta_long <- fit_tbl %>%
  dplyr::select(K, dAIC, dBIC) %>%
  pivot_longer(-K, names_to = "delta", values_to = "value")

ggplot(delta_long, aes(x = K, y = value, group = delta)) +
  geom_hline(yintercept = 0) +
  geom_line() +
  geom_point() +
  facet_wrap(~delta, scales = "free_y") +
  labs(x = "K", y = "Change vs K-1 (negative = better)")
```

We observe that @tbl-fit shows the latent class model estimates for one to five classes, while @fig-visaic presents both the Akaike Information Criterion (AIC) and the Bayesian Information Criterion (BIC) in a figure to aid understanding of the fit criteria.

The AIC shows a constant decreasing across the whole range tested (8003.9 → 7523.2) and therefore does not by itself identify a maximum at K = 5, but the size of the improvement shrinks sharply with each additional class (dAIC = -259.7, -115.1, -72.6, -33.3 for K = 2 through 5): most of the gain in fit is already captured by K = 3-4, and each class added beyond that buys comparatively little. On the other hand, the BIC is minimized at K = 4 (7722.1), after falling steadily from K = 1 (8041.7) through K = 4 and then rising again at K = 5 (7731.4, dBIC = +9.3), visualizing a U-shape that marks the BIC-optimal model. Taken together, the BIC minimum together with the diminishing AIC returns favor the 4-class model as the most parsimonious solution.



```{r}
#| label: tbl-entropy
#| tbl-cap: "Classification quality by number of classes"
#| tbl-cap-location: top
#| include: false

entropy_tbl <- bind_rows(
  lapply(seq_along(fits), function(i) {
    K <- get_K(fits[[i]])
    if (K == 1) return(tibble(K = 1, entropy = NA, mean_max_p = NA,
                               p10_max_p = NA, avg_diag_min = NA, avg_diag_mean = NA))
    entropy_stats(fits[[i]]) %>% mutate(K = K, .before = 1)
  })
)

entropy_tbl %>%
  mutate(across(-K, ~round(., 3))) %>%
  kableExtra::kable(format = "html",
                     align = "c",
                     booktabs = TRUE,
                     caption = NULL) %>%
  kableExtra::kable_styling(full_width = TRUE,
                             latex_options = "hold_position",
                             bootstrap_options = c("striped", "bordered", "condensed"),
                             font_size = 23)
```

```{r}
#| label: fig-entropy
#| out-width: '80%'
#| fig-asp: 0.6
#| fig-cap: "Relative entropy by K (higher = better class separation)"
#| fig-cap-location: top

ggplot(filter(entropy_tbl, K > 1), aes(x = K, y = entropy)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = entropy_tbl$K) +
  labs(x = "Number of classes (K)", y = "Relative entropy")
```


To support this result, we estimated the relative entropy for the class solutions previously tested. @fig-entropy shows the relative entropy rises from 0.657 (K=2) to 0.716 (K=3) to 0.809 (K=4) — crossing the conventional 0.8 threshold usually taken as "good" separation — and then falls back to 0.782 at K=5. These results suggest adding a fifth class does not sharpen the typology further; it makes assignment noisier, which is consistent with K=5 overfitting the data (as already suggested by its shrinking AIC gains and its BIC increase). Thus, the 4-class solution combines the best statistical fit with the cleanest, most confidently separated classes, reinforcing its status as the optimal model.

### Four-classes solution

The @tbl-prevalence shows the conditional probabilities for each item associated with a latent class, along with their respective counts. The four-class solution suggests that respondents are not primarily differentiated along a simple continuum from “more meritocratic” to “less meritocratic.” Instead, the latent classes are organized by distinct combinations of: (1) perceptions of meritocracy, (2) perceptions of non-meritocratic advantage, and (3) normative preferences regarding whether merit or privilege should determine success. As such, the results reveal heterogeneity not only in empirical diagnoses of how society works, but also in normative evaluations of whether these mechanisms are considered acceptable. This interpretation is consistent with the item-response profiles across classes.

```{r}

class_colors <- c(
  "1" = "#961010",
  "2" = "#ffbb00",
  "3" = "#1e358a",
  "4" = "#78c99a"
)

fit4 <- fits[[4]]

# 2) Extract conditional probabilities per item and class
# fit4$probs is a list: each element is K x Rj matrix.
# For dichotomous items: K x 2 (categories = colnames or "1","2")
probs_long <- bind_rows(lapply(names(fit4$probs), function(item) {
  mat <- fit4$probs[[item]]
  df <- as.data.frame(mat)
  df$class <- 1:nrow(mat)
  df$item <- item
  df
})) %>%
  pivot_longer(cols = -c(class, item), names_to = "category", values_to = "prob") %>%
  mutate(
    class = as.integer(class),
    prob = as.numeric(prob)
  )

# If levels are c("0","1") or c("1","2") you'll want the "higher"/"yes" one.
# Here I will assume endorsement is the LAST level (most common after dichotomization).
endorsement_level <- tail(levels(db_lca$a1), 1)

# 4) Build a clean table: P(endorsement | class) for each item
p_endorse <- probs_long %>%
  filter(category == "Pr(2)") %>%
  dplyr::select(item, class, p = prob) %>%
  mutate(item = factor(item, levels = names(fit4$probs)))

# 5) Plot: profile lines (one line per class)
p_endorse <- p_endorse %>% 
  mutate(
    variable = case_when(item == "a1" ~ "perc_effort",
                         item == "a2" ~ "perc_talent",
                         item == "a3" ~ "perc_rich_parents",
                         item == "a4" ~ "perc_contacts",
                         item == "a5" ~ "pref_effort",
                         item == "a6" ~ "pref_talent",
                         item == "a7" ~ "pref_rich_parents",
                         item == "a8" ~ "pref_contacts"
                         ),
    variable = factor(variable,
                      levels = c("perc_effort",
                                 "perc_talent",
                                 "perc_rich_parents",
                                 "perc_contacts",
                                 "pref_effort",
                                 "pref_talent",
                                 "pref_rich_parents",
                                 "pref_contacts")))
```

```{r}
#| label: fig-prob
#| out-width: '100%'
#| fig-asp: 0.8
#| fig-cap: "P(endorsement | class) for each item"
#| fig-cap-location: top
#| warning: false

p_endorse <- p_endorse %>% 
  mutate(class = factor(class),
         p = round(p,2))

p_endorse_reordered <- p_endorse %>%
  mutate(class = as.character(class)) %>%
  mutate(class = recode(class,
                        "3" = "1",
                        "4" = "2",
                        "2" = "3",
                        "1" = "4")) %>%
  mutate(class = factor(class, levels = c("1","2","3","4")))

g_class <- ggplot(p_endorse_reordered, aes(x = variable, y = p, group = class, colour = class)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +  
  scale_color_manual(values = class_colors) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Item", y = paste0("P(Y = ", "High", " | class)"), group = "Class", colour = "Class") +
  theme(legend.position = "top",
        text = element_text(size = 14),
        axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1))



ggplotly(g_class)
```

```{r}
#| label: tbl-prevalence
#| tbl-cap: "Items x classes with probabilities"
#| tbl-cap-location: top

# 7) Report table (wide): items x classes with probabilities
# Classes recoded/ordered the same way as in fig-prob, so columns line up
# with the legend order shown there.
class_prev <- tibble(
  class = 1:length(fit4$P),
  pi = as.numeric(fit4$P)
) %>%
  mutate(pi_pct = 100 * pi) %>%
  mutate(class = as.character(class)) %>%
  mutate(class = recode(class,
                        "3" = "1",
                        "4" = "2",
                        "2" = "3",
                        "1" = "4")) %>%
  mutate(class = factor(class, levels = c("1","2","3","4"))) %>%
  arrange(class)

report_probs <- p_endorse_reordered %>%
  mutate(class = paste0("Class_", as.character(class))) %>%
  pivot_wider(names_from = class, values_from = p) %>%
  mutate(across(starts_with("Class_"), ~ round(.x, 3))) %>%
  dplyr::select(item, variable, paste0("Class_", 1:4))

prev_lab <- class_prev %>%
  transmute(class = paste0("Class_", as.character(class)),
            lab = paste0(class, " (", round(pi_pct, 1), "%)"))

report_probs2 <- report_probs
for (i in seq_len(nrow(prev_lab))) {
  old <- prev_lab$class[i]
  new <- prev_lab$lab[i]
  names(report_probs2)[names(report_probs2) == old] <- new
}

report_probs2[,-c(1)] %>% 
   kableExtra::kable(format = "html",
                    align = "c",
                    booktabs = T,
                    escape = F,
                    caption = NULL) %>%
  kableExtra::kable_styling(full_width = T,
                            latex_options = "hold_position",
                            bootstrap_options=c("striped", "bordered", "condensed"),
                            font_size = 23)
```

**Class 1 (%15.1)**

The first class combines a medium-level perception of meritocracy with a high perception of privilege's influence. Its members believe that success does not depend on effort and talent, but rather on factors such as social background and networks. However, at the normative level, they strongly support meritocracy, especially the effort, while rejecting the legitimacy of privilege advantages.

This class is particularly important because it captures a configuration in which meritocracy is upheld as a moral ideal while privilege is recognized as an empirical reality. In this sense, respondents in this group clearly distinguish between how society works and how it should work. It does not consider the current system meritocratic, but it strongly advocates that it be so.


**Class 2 (%19.6)**

The second class follows a similar descriptive framework to the first, showing high levels of privilege indicators but low levels of meritocratic functions, with effort almost imperceptible for this group. However, the substantial difference is that this class accepts that society operates according to norms based on family background and social capital.

In this way, this profile combines a realistic diagnosis of a socially stratified system with a normative legitimation of privilege. These respondents do not merely recognize that success is shaped by inherited or relational advantages; they also appear to accept that this is appropriate, without overlooking the role of effort. Importantly, their endorsement of meritocratic rewards does not operate as a moral alternative to privilege, but rather coexists with it.


**Class 3 (%44.5)**

The third class displays high levels of both perceived meritocracy and perceived non-meritocracy. Respondents in this group strongly believe that effort and talent are rewarded in society, while also recognizing the importance of wealthy parents and social contacts. Normatively, they express very strong support for effort as the predominant factor of rewards distribution, but only moderate acceptance of privilege advantages.

This class suggests a belief system in which merit and privilege are understood as operating simultaneously. These respondents do not deny the role of structural advantage, yet neither do they reject the view that effort and talent matter. In the normative dimension, they strongly endorse effort, whereas talent receives less consideration; both privilege indicators are supported as legitimate mechanisms for rewards. Compared to Class 2, this group shows a similar willingness to legitimize both effort and merit-based advantage, but it still perceives that meritocratic factors operate actively in society.


**Class 4 (%20.7)**

The fourth class perceives high levels of meritocracy in society, while it views the influence of family background and connections on success as low. Regarding their preferences, people in this group strongly favor effort as the criterion for the distribution of rewards in society, while talent ranks low on their list of preferences, on par with factors of privilege.

This profile reflects a worldview that denies the role of hereditary factors in life trajectories, holding that only individual characteristics such as effort and talent enable one to achieve success. Consistent with their perceptions, the respondents’ normative preferences show support only for effort, drawing a striking distinction between indicators of talent and privilege. Thus, this class holds a naive view of social reality, insisting that effort is the sole criterion governing social stratification, even though it is well known that hereditary factors play a fundamental role.

**Summary**

Taken together, the latent class solution suggests that meritocratic beliefs are best understood as multidimensional and internally heterogeneous. The main contrast is not between those who believe in meritocracy and those who do not, but between different ways of combining empirical perceptions of inequality with normative judgments about its legitimacy. Put differently, the classes vary across two central axes: first, whether respondents perceive success as driven by merit or by privilege; and second, whether they consider these mechanisms normatively acceptable.

This framework helps clarify the broader structure of the results. Class 1 exhibits one of the most intuitive patterns: a perception of moderate meritocracy and a strong influence of privilege, leading to a preference for a greater role for effort and talent and a lesser role for privilege. Class 2 shows only a slight influence of merit but demonstrates that a preference for effort and privilege can coexist, with both factors viewed as legitimate. Class 3 is the most numerous in the population and exhibits, on the one hand, a structural understanding of how society functions, recognizing that both merit and privilege play a role; but, like class two, they consider it legitimate for effort and privilege to contribute to success. Class 4 shows a predominance of effort in both the descriptive and normative dimensions, revealing a certain degree of naivety among respondents in failing to recognize the role that hereditary circumstances play in society. Overall, the results show that support for meritocracy does not necessarily imply denial of privilege, and that recognition of privilege does not necessarily entail its rejection, thereby enabling us to better understand how beliefs about meritocracy and privilege take shape in people's minds.
