---
bibliography: "../input/bib/merit-factorial.bib"
csl: "../input/bib/apa6.csl"
editor_options: 
  chunk_output_type: console
---


```{r}
#| echo: false
#| output: false
#| include: false

library(tidyverse)
library(here)

load(file = here("output", "data", "db_long_proc.RData"))


db_long %>% 
  na.omit() %>% 
  group_by(ola) %>% 
  tally()

db_long %>% 
  na.omit() %>% 
  group_by(ola) %>% 
  summarise(m = mean(age),
            sd = sd(age))

db_long %>% 
  na.omit() %>% 
  group_by(ola, p20) %>% 
  tally()

db_long %>% 
  na.omit() %>% 
  group_by(cohort_level) %>% 
  tally()

db_long %>% 
  na.omit() %>% 
  group_by(cohort_level) %>% 
  summarise(m = mean(age),
            sd = sd(age))

db_long %>% 
  na.omit() %>% 
  group_by(cohort_level, p20) %>% 
  tally()

```

# Method

## Participants

This study uses secondary data from the 2023 and 2024 waves of the Panel Survey on Education and Meritocracy (EDUMER), which examines school-age students’ beliefs, attitudes, and behaviors regarding meritocracy, inequality, and citizenship[^1]. The questionnaire was administered to sixth-grade and first-year secondary students in nine schools in Chile’s Metropolitan and Valparaíso regions. Although the sample was non-probabilistic and lacked quota controls, a minimum target of 900 students was set to ensure adequate statistical power. After data processing and listwise deletion, the analytical sample included 846 students in Wave 1 (386 girls, 421 boys, 39 identifying as other; $M_{age}$ = 13.4, $SD_{age}$ = 1.6) and 662 in Wave 2 (303 girls, 338 boys, 21 identifying as other; $M_{age}$ = 14.4, $SD_{age}$ = 1.6), yielding a retention rate of 78.3% (attrition = 21.7%). The main sources of attrition were school transfers between academic years and high absenteeism in several schools.



[^1]: The survey was collected within the research project “Meritocracy in schools: Moral foundations of the education market and its implications for civic education in Chile” (FONDECYT Regular 2021 Nº 1210847).

## Procedure

Data were collected by a professional research firm using online CAWI questionnaires. All participants received informed consent, which was reviewed and approved by a parent or legal guardian. To enable longitudinal linkage while protecting confidentiality, the survey firm generated anonymous unique identifiers for each student, and no directly identifying information was included in the analytic files.


## Measures

### Scale of perceptions and preferences of meritocracy

Meritocratic and non-meritocratic perceptions and preferences were measured using the same items proposed in the original scale [@castillo_multidimensional_2023]. Each of the four dimensions of the scale is measured by two items. The dimensions are:

- Perceived meritocracy: the extent to which effort and ability are rewarded in Chile,

- Perceived non-meritocracy: the extent to which success is perceived as linked to connections and family wealth,

- Preference for meritocracy: agreement with that those who work harder or are more talented should be better rewarded,

- Preference for non-meritocracy: agreement with that it is acceptable for individuals with better connections or wealthy parents to achieve greater success  (see @tbl-merit).
 
Each item is rated on a four-point Likert scale ranging from “strongly disagree” (1) to “strongly agree” (4), therefore higher scores indicate stronger endorsement of the corresponding perception or preference.

```{r}
#| tbl-cap: Items for perceptions and preferences for meritocracy scale
#| tbl-cap-location: top
#| echo: false
#| label: tbl-merit

 # remotes::install_github('haozhu233/kableExtra@a6af5c0')  # for problems with collapse_rows()
 pacman::p_load(knitr, kableExtra, dplyr)
 table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }
 tabitems <- read.csv(file = "../input/tables/table-01.csv",header = 1,sep = ",",encoding = "UTF-8") # call file  generated externally for the table
 cnames <- c("Components", "Dimensions","Item (English)" , "Item original (Spanish)")

 knitr::kable(tabitems, "latex", booktabs = T, linesep = "",col.names = cnames) %>%
   kableExtra::kable_styling(
     full_width = T,
     latex_options = c("hold_position"),
     position = "center",
     font_size = 9.5,
     bootstrap_options=c("striped", "bordered")) %>%
   kableExtra::column_spec(column = 1, width = "2cm", ) %>%
   kableExtra::column_spec(column = 2,width = "2 cm") %>%
   kableExtra::collapse_rows(columns = 1:2,valign = "middle")
```



### Cohort level

To distinguish students’ academic level across waves for conditional and multi-group invariance analyses, we created a cohort indicator capturing whether the respondent belonged to the primary (sixth grade) or secondary (first-year secondary) cohort at the time of each survey wave. Descriptive statistics[^2] by cohort are reported in @tbl-cohort. 

[^2]: @tbl-cohort totals (N = 1,508) correspond to the sum of person-wave observations (846 at T1 + 662 at T2), rather than unique individuals.


```{r}
#| tbl-cap: Descriptive statistics for cohort level
#| tbl-cap-location: top
#| echo: false
#| label: tbl-cohort
#| message: false
#| warning: false


tot <- db_long %>% 
  na.omit() %>% 
  group_by(ola, cohort_level) %>% 
  tally() %>% 
  rename(n_total = n) %>% 
  mutate(ola = if_else(ola == "1", "Wave 1", "Wave 2")) %>% 
  ungroup()

tot <- tot %>% 
    add_row(
      ola = "Wave 1",
      cohort_level = "Total",
      n_total = sum(tot[c(1,2),3])
    ) %>% 
  add_row(
      ola = "Wave 2",
      cohort_level = "Total",
      n_total = sum(tot[c(3,4),3]))


edad <- db_long %>% 
  na.omit() %>% 
  group_by(ola, cohort_level) %>% 
  summarise(m = round(mean(age),2),
            sd = round(sd(age),2),
            .groups = "drop_last") %>% 
  mutate(ola = if_else(ola == "1", "Wave 1", "Wave 2")) %>% 
  ungroup()

edad_ola <- db_long %>% 
  na.omit() %>% 
  group_by(ola) %>% 
  summarise(m = round(mean(age),2),
            sd = round(sd(age),2),
            .groups = "drop_last") %>% 
  mutate(ola = if_else(ola == "1", "Wave 1", "Wave 2"),
         cohort_level = "Total") %>% 
  select(ola, cohort_level, m, sd)


edad <- bind_rows(edad, edad_ola)

gen <- db_long %>% 
  na.omit() %>% 
  group_by(ola, cohort_level, p20) %>% 
  tally() %>% 
  ungroup() %>% 
  pivot_wider(id_cols = c(ola, cohort_level),
              names_from = p20,
              values_from = n) %>% 
  mutate(ola = if_else(ola == "1", "Wave 1", "Wave 2"))

gen_ola <- db_long %>% 
  na.omit() %>% 
  group_by(ola, p20) %>% 
  tally() %>% 
  ungroup() %>% 
  pivot_wider(id_cols = c(ola),
              names_from = p20,
              values_from = n) %>% 
  mutate(ola = if_else(ola == "1", "Wave 1", "Wave 2"),
         cohort_level = "Total") %>% 
  select(ola, cohort_level, H, M, O)

gen <- bind_rows(gen, gen_ola)

left_join(tot, edad) %>% 
  left_join(., gen) %>% 
  arrange(ola) %>% 
   knitr::kable(., "latex", escape = T,
                booktabs = T, 
                linesep = "",
                col.names = c("Wave", "Cohort level", "N", "Mean", "SD", "Man", "Women", "Others")) %>%
    kableExtra::add_header_above(c(" " = 3, "Age" = 2, "Gender" = 3)) %>% 
   kableExtra::kable_styling(
     full_width = T,
     latex_options = c("hold_position"),
     position = "center",
     font_size = 9.5,
     bootstrap_options=c("striped", "bordered")) %>% 
  kableExtra::collapse_rows(1)

```




## Analytical strategy

### Measurement model and estimation

To evaluate the underlying structure of the scale, we employed Confirmatory Factor Analysis (CFA) based on a measurement model with four latent factors, using Diagonally Weighted Least Squares with robust correction (WLSMV) estimation. This estimator is particularly suitable for ordinal data, such as four-point Likert-type scales, as it avoids the bias associated with treating categorical data as continuous [@kline_principles_2023].

Model fit was assessed following the guidelines of Brown [-@brown_confirmatory_2015], using several indices: the Comparative Fit Index (CFI) and the Tucker-Lewis Index (TLI), with acceptable values above 0.95; the Root Mean Square Error of Approximation (RMSEA), with values below 0.06 indicating good fit; and the Chi-square statistic (acceptable fit indicated by $p$ > 0.05 and a Chi-square/df ratio < 3).

### Measurement invariance

A key contribution of this study is its assessment of measurement stability through factorial invariance testing [@davidov_measurement_2014]. We implemented two complementary strategies: cross-group invariance across school levels (primary vs. secondary students) and longitudinal invariance across two panel waves. Given the ordinal nature of the items, we used estimation procedures appropriate for ordered categorical indicators, following @liu_testing_2017 for longitudinal analyses and @svetina_multiplegroup_2020 for cross-group tests. We assessed invariance using a hierarchical sequence of nested confirmatory factor analysis models that imposed progressively stricter equality constraints. For longitudinal invariance, we estimated a configural model specifying the same factor structure over time while allowing item parameters to vary, followed by metric, scalar, and strict models constraining factor loadings, thresholds/intercepts, and residual variances, respectively. For cross-group invariance, we estimated a configural model, then a threshold-invariant model, and finally a model jointly constraining thresholds and factor loadings across groups.

In addition to the chi-square difference test, we assessed invariance using changes in approximate fit indices. Following @chen_sensitivity_2007, we treated $\Delta$CFI as the primary criterion and $\Delta$RMSEA as supplementary, indicating noninvariance when $\Delta$CFI $\leq$ −0.010 and $\Delta$RMSEA $\geq$ 0.015.


All analyses were performed using the *lavaan* package in R version 4.5.2. The hypotheses of this research were [pre-registered](https://osf.io/hjat2/) on the Open Science Framework (OSF).
