---
bibliography: "../input/bib/merit-factorial.bib"
csl: "../input/bib/apa6.csl"
editor_options: 
  chunk_output_type: console
---

## Invariance analysis

```{r}
#| echo: false
#| message: false
#| warning: false

library(knitr)

knitr::opts_chunk$set(echo = F,
                      warning = F,
                      error = F, 
                      message = F) 

table_format <- if (is_html_output()) {
  "html"
} else if (is_latex_output()) {
  "latex"
}
table_format2 <- if (is_html_output()) {
  T
} else if (is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```

```{r}
#| include: false
#| echo: false

if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,
  sjmisc,
  sjPlot,
  here,
  lavaan,
  psych,
  corrplot,
  ggdist,
  patchwork,
  semTools,
  gtools,
  kableExtra,
  scales
)

options(scipen = 999)
rm(list = ls())
```

```{r}
#| echo: false
#| output: false

load(file = here("output", "data", "db_long_proc.RData"))

names(db_long)
glimpse(db_long)
dim(db_long)
```

```{r}
#| label: data 2
#| results: asis
#| echo: false
#| warning: false
#| output: false

# Crear base en formato wide con dummy de cohorte (0 = primaria, 1 = secundaria)
db_invariance <- db_long %>%
  group_by(id_estudiante) %>%
  mutate(
    cohort_level = first(cohort_level),
    cohort_dummy = case_when(
      cohort_level == "Primary" ~ 0,
      cohort_level == "Secondary" ~ 1,
      TRUE ~ NA_real_
    )
  ) %>%
  ungroup() %>%
  select(id_estudiante, cohort_dummy, ola, starts_with(c("perc", "pref"))) %>%
  pivot_wider(
    id_cols = c(id_estudiante, cohort_dummy),
    names_from = ola,
    names_glue = "{.value}{ola}",
    values_from = c(
      perc_effort, perc_talent,
      perc_rich_parents, perc_contact,
      pref_effort, pref_talent,
      pref_rich_parents, pref_contact
    )
  ) %>%
  na.omit() %>%
  rename_with(~ str_replace_all(., "_", ""))

# Reescalar variables y convertir a ordinales
db_invariance <- db_invariance %>% 
  mutate(
    across(
      .cols = -c(idestudiante, cohortdummy),
      .fns = ~ case_when(. == 1 ~ 0,
                         . == 2 ~ 1,
                         . == 3 ~ 2,
                         . == 4 ~ 3)
    )
  ) %>%
  mutate(
    across(
      .cols = -c(idestudiante),
      .fns = ~ ordered(.)
    )
  )
```

```{r}
#| label: conf-model
#| tbl-cap: "Longitudinal Invariance Test"
#| results: asis
#| echo: false
#| warning: false

# First, define the configural model, using the repeated measures factors and
# indicators
baseline_model_smt <- ('

###########################################
# Definición de factores (1 marcador por factor)
###########################################

percmerit1  =~ 1*perceffort1     + perctalent1
percmerit2  =~ 1*perceffort2     + perctalent2

percnmerit1 =~ 1*percrichparents1 + perccontact1
percnmerit2 =~ 1*percrichparents2 + perccontact2

prefmerit1  =~ 1*prefeffort1     + preftalent1
prefmerit2  =~ 1*prefeffort2     + preftalent2

prefnmerit1 =~ 1*prefrichparents1 + prefcontact1
prefnmerit2 =~ 1*prefrichparents2 + prefcontact2


###########################################
# Covarianzas entre errores (mismo ítem, distintas olas)
###########################################

perceffort1 ~~ perceffort2
perctalent1 ~~ perctalent2

percrichparents1 ~~ percrichparents2
perccontact1     ~~ perccontact2

prefeffort1 ~~ prefeffort2
preftalent1 ~~ preftalent2

prefrichparents1 ~~ prefrichparents2
prefcontact1     ~~ prefcontact2
')



# Model Estimation
fit_baseline <- cfa(baseline_model_smt, data = db_invariance,
                    ordered = c("perceffort1", "perctalent1", "perceffort2", "perctalent2",
                                "percrichparents1", "perccontact1", "percrichparents2", "perccontact2",
                                "prefeffort1", "preftalent1", "prefeffort2", "preftalent2",
                                "prefrichparents1", "prefcontact1", "prefrichparents2", "prefcontact2"),
                    parameterization = "theta",
                    estimator = "WLSMV")
```

```{r}
#| label: Weak-Model
#| tbl-cap: "Longitudinal Invariance Weak Model"
#| results: asis
#| echo: false
#| warning: false

loadinginv_model_smt <- ('

# Igualación de cargas (invarianza débil)

percmerit1  =~ 1*perceffort1 + pe_loading*perctalent1
percmerit2  =~ 1*perceffort2 + pe_loading*perctalent2

percnmerit1 =~ 1*percrichparents1 + prc_loading*perccontact1
percnmerit2 =~ 1*percrichparents2 + prc_loading*perccontact2

prefmerit1  =~ 1*prefeffort1 + pfe_loading*preftalent1
prefmerit2  =~ 1*prefeffort2 + pfe_loading*preftalent2

prefnmerit1 =~ 1*prefrichparents1 + pfrc_loading*prefcontact1
prefnmerit2 =~ 1*prefrichparents2 + pfrc_loading*prefcontact2


# Varianzas y covarianzas latentes

percmerit1  ~~ percmerit1 + percmerit2
percmerit2  ~~ percmerit2
percnmerit1 ~~ percnmerit1 + percnmerit2
percnmerit2 ~~ percnmerit2
prefmerit1  ~~ prefmerit1 + prefmerit2
prefmerit2  ~~ prefmerit2
prefnmerit1 ~~ prefnmerit1 + prefnmerit2
prefnmerit2 ~~ prefnmerit2


# Umbrales (uno igualado por ítem)

# perceffort1 | pe1t1*t1 + pe1t2*t2
# perceffort2 | pe2t1*t1 + pe2t2*t2
# perctalent1 | pt1t1*t1 + pt1t2*t2
# perctalent2 | pt2t1*t1 + pt2t2*t2
# percrichparents1 | prp1t1*t1 + prp1t2*t2
# percrichparents2 | prp2t1*t1 + prp2t2*t2
# perccontact1 | pc1t1*t1 + pc1t2*t2
# perccontact2 | pc2t1*t1 + pc2t2*t2
# prefeffort1 | pre1t1*t1 + pre1t2*t2
# prefeffort2 | pre2t1*t1 + pre2t2*t2
# preftalent1 | prt1t1*t1 + prt1t2*t2
# preftalent2 | prt2t1*t1 + prt2t2*t2
# prefrichparents1 | pfrp1t1*t1 + pfrp1t2*t2
# prefrichparents2 | pfrp2t1*t1 + pfrp2t2*t2
# prefcontact1 | pfc1t1*t1 + pfc1t2*t2
# prefcontact2 | pfc2t1*t1 + pfc2t2*t2
 

# Covarianzas entre errores

perceffort1 ~~ perceffort2
perctalent1 ~~ perctalent2
percrichparents1 ~~ percrichparents2
perccontact1 ~~ perccontact2
prefeffort1 ~~ prefeffort2
preftalent1 ~~ preftalent2
prefrichparents1 ~~ prefrichparents2
prefcontact1 ~~ prefcontact2
')


fit_loadinginv <- cfa(loadinginv_model_smt, data = db_invariance,
                      ordered = c("perceffort1", "perctalent1", "perceffort2", "perctalent2",
                                  "percrichparents1", "perccontact1", "percrichparents2", "perccontact2",
                                  "prefeffort1", "preftalent1", "prefeffort2", "preftalent2",
                                  "prefrichparents1", "prefcontact1", "prefrichparents2", "prefcontact2"),
                      parameterization = "theta",
                      estimator = "WLSMV")
```

```{r}
#| label: strong-model
#| tbl-cap: "Longitudinal Invariance Strong Model"
#| results: asis
#| echo: false
#| warning: false

thresholdinv_model_smt <- ('

# Igualación de cargas (como en el modelo débil)

percmerit1  =~ 1*perceffort1 + pe_loading*perctalent1
percmerit2  =~ 1*perceffort2 + pe_loading*perctalent2

percnmerit1 =~ 1*percrichparents1 + prc_loading*perccontact1
percnmerit2 =~ 1*percrichparents2 + prc_loading*perccontact2

prefmerit1  =~ 1*prefeffort1 + pfe_loading*preftalent1
prefmerit2  =~ 1*prefeffort2 + pfe_loading*preftalent2

prefnmerit1 =~ 1*prefrichparents1 + pfrc_loading*prefcontact1
prefnmerit2 =~ 1*prefrichparents2 + pfrc_loading*prefcontact2


# Varianzas y covarianzas latentes

percmerit1  ~~ percmerit1 + percmerit2
percmerit2  ~~ percmerit2
percnmerit1 ~~ percnmerit1 + percnmerit2
percnmerit2 ~~ percnmerit2
prefmerit1  ~~ prefmerit1 + prefmerit2
prefmerit2  ~~ prefmerit2
prefnmerit1 ~~ prefnmerit1 + prefnmerit2
prefnmerit2 ~~ prefnmerit2


# Thresholds iguales entre olas (invarianza fuerte)

perceffort1     | pe_t1*t1 + pe_t2*t2
perceffort2     | pe_t1*t1 + pe_t2*t2
perctalent1     | pt_t1*t1 + pt_t2*t2
perctalent2     | pt_t1*t1 + pt_t2*t2
percrichparents1| prp_t1*t1 + prp_t2*t2
percrichparents2| prp_t1*t1 + prp_t2*t2
perccontact1    | pc_t1*t1 + pc_t2*t2
perccontact2    | pc_t1*t1 + pc_t2*t2
prefeffort1     | pre_t1*t1 + pre_t2*t2
prefeffort2     | pre_t1*t1 + pre_t2*t2
preftalent1     | prt_t1*t1 + prt_t2*t2
preftalent2     | prt_t1*t1 + prt_t2*t2
prefrichparents1| pfrp_t1*t1 + pfrp_t2*t2
prefrichparents2| pfrp_t1*t1 + pfrp_t2*t2
prefcontact1    | pfc_t1*t1 + pfc_t2*t2
prefcontact2    | pfc_t1*t1 + pfc_t2*t2


# Interceptos fijos

perceffort1 + perctalent1 ~ 0*1
perceffort2 + perctalent2 ~ 0*1
percrichparents1 + perccontact1 ~ 0*1
percrichparents2 + perccontact2 ~ 0*1
prefeffort1 + preftalent1 ~ 0*1
prefeffort2 + preftalent2 ~ 0*1
prefrichparents1 + prefcontact1 ~ 0*1
prefrichparents2 + prefcontact2 ~ 0*1


# Varianzas únicas

perceffort1 ~~ 1*perceffort1
perctalent1 ~~ 1*perctalent1
percrichparents1 ~~ 1*percrichparents1
perccontact1 ~~ 1*perccontact1
prefeffort1 ~~ 1*prefeffort1
preftalent1 ~~ 1*preftalent1
prefrichparents1 ~~ 1*prefrichparents1
prefcontact1 ~~ 1*prefcontact1

perceffort2 ~~ NA*perceffort2
perctalent2 ~~ NA*perctalent2
percrichparents2 ~~ NA*percrichparents2
perccontact2 ~~ NA*perccontact2
prefeffort2 ~~ NA*prefeffort2
preftalent2 ~~ NA*preftalent2
prefrichparents2 ~~ NA*prefrichparents2
prefcontact2 ~~ NA*prefcontact2


# Covarianzas entre errores longitudinales

perceffort1 ~~ perceffort2
perctalent1 ~~ perctalent2
percrichparents1 ~~ percrichparents2
perccontact1 ~~ perccontact2
prefeffort1 ~~ prefeffort2
preftalent1 ~~ preftalent2
prefrichparents1 ~~ prefrichparents2
prefcontact1 ~~ prefcontact2
')

fit_thresholdinv <- cfa(thresholdinv_model_smt, data = db_invariance,
                        ordered = c("perceffort1", "perctalent1", "perceffort2", "perctalent2",
                                    "percrichparents1", "perccontact1", "percrichparents2", "perccontact2",
                                    "prefeffort1", "preftalent1", "prefeffort2", "preftalent2",
                                    "prefrichparents1", "prefcontact1", "prefrichparents2", "prefcontact2"),
                        parameterization = "theta",
                        estimator = "WLSMV")
```

```{r}
#| label: strict-model
#| tbl-cap: "Longitudinal Invariance Strict Model"
#| results: asis
#| echo: false
#| warning: false

uniquenessinv_model_smt <- ('

# Cargas factoriales (invarianza débil)

percmerit1  =~ 1*perceffort1 + pe_loading*perctalent1
percmerit2  =~ 1*perceffort2 + pe_loading*perctalent2

percnmerit1 =~ 1*percrichparents1 + prc_loading*perccontact1
percnmerit2 =~ 1*percrichparents2 + prc_loading*perccontact2

prefmerit1  =~ 1*prefeffort1 + pfe_loading*preftalent1
prefmerit2  =~ 1*prefeffort2 + pfe_loading*preftalent2

prefnmerit1 =~ 1*prefrichparents1 + pfrc_loading*prefcontact1
prefnmerit2 =~ 1*prefrichparents2 + pfrc_loading*prefcontact2


# Varianzas y covarianzas de factores latentes

percmerit1  ~~ percmerit1 + percmerit2
percmerit2  ~~ percmerit2
percnmerit1 ~~ percnmerit1 + percnmerit2
percnmerit2 ~~ percnmerit2
prefmerit1  ~~ prefmerit1 + prefmerit2
prefmerit2  ~~ prefmerit2
prefnmerit1 ~~ prefnmerit1 + prefnmerit2
prefnmerit2 ~~ prefnmerit2


# Medias latentes

percmerit1  ~ 0*1
percmerit2  ~ 1
percnmerit1 ~ 0*1
percnmerit2 ~ 1
prefmerit1  ~ 0*1
prefmerit2  ~ 1
prefnmerit1 ~ 0*1
prefnmerit2 ~ 1


# Umbrales (invarianza fuerte)

perceffort1     | pe_t1*t1 + pe_t2*t2
perceffort2     | pe_t1*t1 + pe_t2*t2
perctalent1     | pt_t1*t1 + pt_t2*t2
perctalent2     | pt_t1*t1 + pt_t2*t2
percrichparents1| prp_t1*t1 + prp_t2*t2
percrichparents2| prp_t1*t1 + prp_t2*t2
perccontact1    | pc_t1*t1 + pc_t2*t2
perccontact2    | pc_t1*t1 + pc_t2*t2
prefeffort1     | pre_t1*t1 + pre_t2*t2
prefeffort2     | pre_t1*t1 + pre_t2*t2
preftalent1     | prt_t1*t1 + prt_t2*t2
preftalent2     | prt_t1*t1 + prt_t2*t2
prefrichparents1| pfrp_t1*t1 + pfrp_t2*t2
prefrichparents2| pfrp_t1*t1 + pfrp_t2*t2
prefcontact1    | pfc_t1*t1 + pfc_t2*t2
prefcontact2    | pfc_t1*t1 + pfc_t2*t2


# Interceptos fijos

perceffort1 + perctalent1 ~ 0*1
perceffort2 + perctalent2 ~ 0*1
percrichparents1 + perccontact1 ~ 0*1
percrichparents2 + perccontact2 ~ 0*1
prefeffort1 + preftalent1 ~ 0*1
prefeffort2 + preftalent2 ~ 0*1
prefrichparents1 + prefcontact1 ~ 0*1
prefrichparents2 + prefcontact2 ~ 0*1


# Varianzas únicas (invarianza estricta)
# Fijadas a 1 en todas las olas

perceffort1 ~~ 1*perceffort1
perceffort2 ~~ 1*perceffort2
perctalent1 ~~ 1*perctalent1
perctalent2 ~~ 1*perctalent2

percrichparents1 ~~ 1*percrichparents1
percrichparents2 ~~ 1*percrichparents2
perccontact1 ~~ 1*perccontact1
perccontact2 ~~ 1*perccontact2

prefeffort1 ~~ 1*prefeffort1
prefeffort2 ~~ 1*prefeffort2
preftalent1 ~~ 1*preftalent1
preftalent2 ~~ 1*preftalent2

prefrichparents1 ~~ 1*prefrichparents1
prefrichparents2 ~~ 1*prefrichparents2
prefcontact1 ~~ 1*prefcontact1
prefcontact2 ~~ 1*prefcontact2


# Covarianzas entre errores (mismo ítem, diferente ola)

perceffort1 ~~ perceffort2
perctalent1 ~~ perctalent2
percrichparents1 ~~ percrichparents2
perccontact1 ~~ perccontact2
prefeffort1 ~~ prefeffort2
preftalent1 ~~ preftalent2
prefrichparents1 ~~ prefrichparents2
prefcontact1 ~~ prefcontact2
')

fit_uniquenessinv <- cfa(uniquenessinv_model_smt, data = db_invariance,
                         ordered = c("perceffort1", "perctalent1", "perceffort2", "perctalent2",
                                     "percrichparents1", "perccontact1", "percrichparents2", "perccontact2",
                                     "prefeffort1", "preftalent1", "prefeffort2", "preftalent2",
                                     "prefrichparents1", "prefcontact1", "prefrichparents2", "prefcontact2"),
                         parameterization = "theta",
                         estimator = "WLSMV")
```

### Longitudinal invariance

@fig-likerplot1 displays the response distributions for the meritocracy scale, distinguishing perceptions (Panel A) from preferences (Panel B) across waves. Across both waves, a clear and stable gap emerges between how students perceive social rewards to operate and how they think they should operate. In the perception items, agreement is consistently high for contacts, parental wealth, and talent, whereas effort is also widely endorsed but to a lesser extent, indicating that students recognize both meritocratic and non-meritocratic determinants of success and view talent as more influential than effort. In the preference items, by contrast, effort is overwhelmingly endorsed as the legitimate basis for rewards, far exceeding its perceived role in practice. Talent occupies a more ambivalent normative position, parental wealth remains evenly divided, and contacts are more often accepted than rejected, suggesting that non-meritocratic criteria are not uniformly seen as illegitimate. Wave 2 closely replicates these patterns, with only modest shifts. Overall, the descriptive results reveal a persistent paradox: strong endorsement of effort-centered meritocratic ideals alongside skepticism about their realization in practice.






```{r}
#| label: fig-likerplot1
#| fig-align: center
#| fig-cap: "Distribution of responses on the scale of perceptions and preferences regarding meritocracy by wave"
#| fig-cap-location: top
#| fig-asp: 0.9
#| out-width: '100%'
#| echo: false
#| warning: false
#| message: false

library(cowplot)

cols  <- RColorBrewer::brewer.pal(4, "RdBu")
items1 <- c("perc_effort","perc_talent","perc_rich_parents","perc_contact")

db_likert1 <- db_long %>%
  filter(ola == 1) %>%
  select(starts_with(c("perc"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items1),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert1 <- db_likert1 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab1 <- db_likert1 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

a <- ggplot(db_likert1, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab1, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(title = "Perceptions", x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 18, 5, 10))

items2 <- c("pref_effort","pref_talent","pref_rich_parents","pref_contact")

db_likert2 <- db_long %>%
  filter(ola == 1) %>%
  select(starts_with(c("pref"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items2),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert2 <- db_likert2 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab2 <- db_likert2 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

b <- ggplot(db_likert2, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab2, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(title = "Preferences", x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 18, 5, 10))

# ----


db_likert3 <- db_long %>%
  filter(ola == 2) %>%
  select(starts_with(c("perc"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items1),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert3 <- db_likert3 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab3 <- db_likert3 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

c <- ggplot(db_likert3, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab3, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 18, 5, 10))

db_likert4 <- db_long %>%
  filter(ola == 2) %>%
  select(starts_with(c("pref"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items2),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert4 <- db_likert4 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab4 <- db_likert4 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

d <- ggplot(db_likert4, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab4, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 18, 5, 10))
# 4) grilla 2x2
panels <- cowplot::plot_grid(a, b, c, d, ncol = 2, align = "hv", axis = "tblr")

# 5) poner Wave 1 / Wave 2 encima (no consume ancho)
panels_labeled <- cowplot::ggdraw(panels) +
  draw_label("Wave 1", x = 0.03, y = 0.75, angle = 0, hjust = 1.5, vjust = 0.5, size = 11) +
  draw_label("Wave 2", x = 0.03, y = 0.25, angle = 0, hjust = 1.5, vjust = 0.5, size = 11)

# 6) leyenda a mano COMO PANEL (sin extraer grobs)
lik_labels <- c("Strongly disagree", "Disagree", "Agree", "Strongly agree")
df_leg <- tibble(cat = factor(lik_labels, levels = lik_labels))

legend_only <- ggplot(df_leg, aes(x = 2, y = 2, fill = cat)) +
  geom_point(shape = 22, size = 6, color = "black", stroke = 0.3) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), clip = "on") +  # “esconde” los puntos
  scale_fill_manual(values = cols, drop = FALSE) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE, title = NULL)) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.text     = element_text(size = 9),
    legend.key.size = unit(4, "mm"),
    plot.margin     = margin(0, 0, 0, 0)
  )
# 7) caption común como panel
cap <- cowplot::ggdraw() +
  draw_label("Source: own elaboration based on EDUMER Panel Survey (N Wave 1 =  846; N Wave 2 = 662)", x = 0, hjust = 0, size = 9)

# 8) final: todo junto
final <- plot_grid(
  panels_labeled,
  legend_only,
  cap,
  ncol = 1,
  rel_heights = c(1.5, 0.10, 0.06)
)

final +
  theme(plot.margin = margin(t = 5.5, r = 5.5, b = 5.5, l = 50))

```


@tbl-longinv reports the results of the longitudinal invariance tests. Overall, the scale achieved the strongest level of invariance (strict invariance) across the two waves. The strict model showed good fit, $\chi^2$(84) = 130.17, CFI = 0.991, RMSEA = 0.031 (90% CI: 0.020–0.040). Moreover, compared with the strong (scalar) model, the deterioration in fit was negligible: $\Delta \chi^2$(4) = 1.635, $\Delta$CFI = 0.000, and $\Delta$RMSEA = -0.002. These differences fall well within commonly used cutoffs for invariance evaluation [@chen_sensitivity_2007], indicating that constraining residual variances in addition to loadings and thresholds/intercepts is tenable. Substantively, this implies that the item–factor relations are stable over time (metric invariance) and that, for a given level of the latent trait, respondents are expected to endorse the same response categories across waves (scalar invariance), supporting the comparability of scores and latent parameters within students over time.

```{r}
#| tbl-cap: "Longitudinal invariance results"
#| tbl-cap-location: top
#| echo: false
#| warning: false
#| label: tbl-longinv

# Compare fit

an1 <- anova(fit_baseline, fit_loadinginv)
an2 <- anova(fit_loadinginv, fit_thresholdinv)
an3 <- anova(fit_thresholdinv, fit_uniquenessinv)

tab01 <- bind_rows(
  as_tibble(an1)[2,],
  as_tibble(an2)[2,],
  as_tibble(an3)[2,]
) %>%
  select("Chisq", "Df", chisq_diff = `Chisq diff`, df_diff = `Df diff`, pvalue = `Pr(>Chisq)`) %>%
  mutate(
    stars = stars.pval(pvalue),
    chisqt = paste0(round(Chisq, 2), " (", Df, ")"),
    decision = ifelse(pvalue > 0.05, "Accept", "Reject"),
    model = c("Weak", "Strong", "Strict")
  ) %>%
  bind_rows(
    tibble(Chisq = an1$Chisq[1], Df = an1$Df[1], chisq_diff = NA, df_diff = NA,
           pvalue = NA, stars = "", chisqt = paste0(round(an1$Chisq[1], 2), " (", an1$Df[1], ")"),
           decision = "Reference", model = "Configural")
  ) %>%
  select(model, chisqt, chisq_diff, df_diff, pvalue, stars, decision) %>%
  mutate(model = factor(model, levels = c("Configural", "Weak", "Strong", "Strict"))) %>%
  arrange(model)


fit.meas <- dplyr::bind_rows(lavaan::fitmeasures(fit_baseline, output ="matrix")[c("chisq","df","cfi","rmsea","rmsea.ci.lower","rmsea.ci.upper"),],
                             lavaan::fitmeasures(fit_loadinginv,  output ="matrix")[c("chisq","df","cfi","rmsea","rmsea.ci.lower","rmsea.ci.upper"),],
                             lavaan::fitmeasures(fit_thresholdinv, output ="matrix")[c("chisq","df","cfi","rmsea","rmsea.ci.lower","rmsea.ci.upper"),],
                             lavaan::fitmeasures(fit_uniquenessinv, output ="matrix")[c("chisq","df","cfi","rmsea","rmsea.ci.lower","rmsea.ci.upper"),])

fit.meas <- fit.meas %>%
  dplyr::mutate(diff.chi2 = chisq    - lag(chisq,default = dplyr::first(chisq)),
                diff.df   = df       - lag(df,   default = dplyr::first(df)),
                diff.cfi  = cfi      - lag(cfi,  default = dplyr::first(cfi)),
                diff.rmsea   = rmsea - lag(rmsea,default = dplyr::first(rmsea))) %>%
  round(3) %>%
  dplyr::mutate(rmsea.ci=paste0(rmsea," \n ", "(",rmsea.ci.lower,"-",rmsea.ci.upper,")"))

tab.inv <- dplyr::bind_cols(tab01,fit.meas) %>%
  dplyr::select(model,chisqt,cfi,rmsea.ci,diff.chi2,diff.df,diff.cfi,diff.rmsea,stars,decision) %>%
  dplyr::mutate(diff.chi2=paste0(diff.chi2," (",diff.df,") ",stars)) %>%
  dplyr::select(model,chisqt,cfi,rmsea.ci,diff.chi2,diff.cfi,diff.rmsea,decision)


col.nam <- c(
  "Model",
  "$\\chi^2$ (df)",
  "CFI",
  "RMSEA (90\\% CI)",
  "$\\Delta \\chi^2$ ($\\Delta$df)",
  "$\\Delta$CFI",
  "$\\Delta$RMSEA",
  "Decision"
)

tab.inv %>% 
  kableExtra::kable(format = "latex",
                    align = "c",
                    booktabs = T,
                    escape = F,
                    caption = NULL,
                    col.names = col.nam) %>%
  kableExtra::kable_styling(full_width = T,
                            latex_options = "hold_position",
                            bootstrap_options=c("striped", "bordered", "condensed"),
                           font_size = 9.5) %>% 
  kableExtra::column_spec(column = 1, width = "2cm") %>%
   kableExtra::column_spec(column = 4,width = "2 cm")
```

As an additional diagnostic, we used univariate score tests to identify potential sources of localized misfit at each step of the invariance hierarchy. The tests revealed no meaningful violations of the imposed equality constraints, suggesting that model fit was not driven by a small subset of parameters. Global score tests and the smallest p-values at each step are reported in the supplementary materials.

### Cohort invariance

@fig-likerplot2 shows response distributions by cohort, separating perceptions and preferences. In primary education, effort and talent are perceived as similarly rewarded, while contacts and family wealth are also seen as advantageous, though with more disagreement. In secondary education, perceptions become less meritocratic: agreement that effort is rewarded declines, whereas contacts and family wealth are seen as more strongly shaping success. Preferences are more stable across cohorts. In both groups, effort is the most strongly endorsed legitimate basis for reward, while talent, family wealth, and especially contacts remain more ambivalent. Notably, contacts are more often accepted than rejected in both cohorts. Overall, the cohort comparison reveals a widening gap between an effort-centered meritocratic ideal and the perceived importance of non-meritocratic advantages in practice.

```{r}
#| label: fig-likerplot2
#| fig-align: center
#| fig-cap: "Distribution of responses on the scale of perceptions and preferences regarding meritocracy by cohort"
#| fig-cap-location: top
#| fig-asp: 0.9
#| out-width: '100%'
#| echo: false
#| warning: false
#| message: false
cols  <- RColorBrewer::brewer.pal(4, "RdBu")
items1 <- c("perc_effort","perc_talent","perc_rich_parents","perc_contact")

db_likert1 <- db_long %>%
  filter(cohort_level == "Primary") %>%
  select(starts_with(c("perc"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items1),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert1 <- db_likert1 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab1 <- db_likert1 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

a <- ggplot(db_likert1, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab1, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(title = "Perceptions", x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 18, 5, 10))

items2 <- c("pref_effort","pref_talent","pref_rich_parents","pref_contact")

db_likert2 <- db_long %>%
  filter(cohort_level == "Primary") %>%
  select(starts_with(c("pref"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items2),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert2 <- db_likert2 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab2 <- db_likert2 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

b <- ggplot(db_likert2, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab2, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(title = "Preferences", x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 18, 5, 10))

# ----


db_likert3 <- db_long %>%
  filter(cohort_level == "Secondary") %>%
  select(starts_with(c("perc"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items1),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert3 <- db_likert3 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab3 <- db_likert3 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

c <- ggplot(db_likert3, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab3, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 18, 5, 10))

db_likert4 <- db_long %>%
  filter(cohort_level == "Secondary") %>%
  select(starts_with(c("pref"))) %>%
  tidyr::drop_na() %>%                     # <- igual que tu na.omit() (complete-cases)
  pivot_longer(everything(), names_to="item", values_to="x") %>%
  mutate(
    item = factor(item, levels = rev(items2),
                  labels = c("Contacts","Rich parents","Talent","Effort")),
    cat  = factor(x, 1:4,
                  c("Strongly disagree","Disagree","Agree","Strongly agree"),
                  ordered = TRUE)
  ) %>% 
  na.omit()

db_likert4 <- db_likert4 %>% count(item, cat) %>% group_by(item) %>% mutate(p = n/sum(n)) %>% ungroup() %>%
  mutate(s = ifelse(as.integer(cat) <= 2, -p, p))

lab4 <- db_likert4 %>%
  mutate(side = ifelse(as.integer(cat) <= 2, "Disagree", "Agree")) %>%
  group_by(item, side) %>% summarise(p = sum(p), .groups="drop") %>%
  mutate(x = ifelse(side=="Disagree", -(p+.03), (p+.03)),
         h = ifelse(side=="Disagree", 1, 0),
         lab = scales::percent(p, 0.1))

d <- ggplot(db_likert4, aes(s, item, fill = cat)) +
  geom_col(width = .8) +
  geom_vline(xintercept = 0, linewidth = .35) +
  geom_text(data = lab4, aes(x = x, y = item, label = lab, hjust = h),
            inherit.aes = FALSE, size = 3) +
  scale_fill_manual(values = cols) +
  scale_x_continuous(labels = \(z) scales::percent(abs(z)), limits = c(-1.1, 1.1)) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme(legend.position = "none", panel.grid.major.y = element_blank(),
        plot.margin = margin(5, 18, 5, 10))
# 4) grilla 2x2
panels <- plot_grid(a, b, c, d, ncol = 2, align = "hv", axis = "tblr")

# 5) poner Wave 1 / Wave 2 encima (no consume ancho)
panels_labeled <- cowplot::ggdraw(panels) +
  draw_label("Primary", x = 0.03, y = 0.75, angle = 0, hjust = 1.2, vjust = 0.5, size = 11) +
  draw_label("Secondary", x = 0.03, y = 0.25, angle = 0, hjust = 1.2, vjust = 0.5, size = 11)

# 6) leyenda a mano COMO PANEL (sin extraer grobs)
lik_labels <- c("Strongly disagree", "Disagree", "Agree", "Strongly agree")
df_leg <- tibble(cat = factor(lik_labels, levels = lik_labels))

legend_only <- ggplot(df_leg, aes(x = 2, y = 2, fill = cat)) +
  geom_point(shape = 22, size = 6, color = "black", stroke = 0.3) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), clip = "on") +  # “esconde” los puntos
  scale_fill_manual(values = cols, drop = FALSE) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE, title = NULL)) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.text     = element_text(size = 9),
    legend.key.size = unit(4, "mm"),
    plot.margin     = margin(0, 0, 0, 0)
  )
# 7) caption común como panel
cap <- cowplot::ggdraw() +
  draw_label("Source: own elaboration based on EDUMER Panel Survey (N Primary =  725; N Secondary = 783)", x = 0, hjust = 0, size = 9)

# 8) final: todo junto
final <- plot_grid(
  panels_labeled,
  legend_only,
  cap,
  ncol = 1,
  rel_heights = c(1.5, 0.10, 0.06)
)

final +
  theme(plot.margin = margin(t = 5.5, r = 5.5, b = 5.5, l = 50))

```



```{r}
#| echo: false
#| include: false


db1 <- db_long %>%
filter(ola == 1) %>% # quedarse con solo la ola 1
select(id_estudiante, starts_with(c("perc", "pref")), cohort_level)

model_restricted <- ('
perc_merit =~ perc_effort + perc_talent
perc_nmerit =~ a*perc_rich_parents + a*perc_contact
pref_merit =~ pref_effort + pref_talent
pref_nmerit =~ pref_rich_parents + pref_contact
perc_effort ~~ pref_talent
')


mbasica_cfa <- cfa(model = model_restricted, 
                   data = subset(db1, cohort_level == "Primary"), 
                   estimator = "WLSMV",
                   ordered = T,
                   std.lv = F)

mmedia_cfa <- cfa(model = model_restricted, 
                  data = subset(db1, cohort_level == "Secondary"), 
                  estimator = "WLSMV",
                  ordered = T,
                  std.lv = F)

summary(mmedia_cfa, standardized = TRUE, fit.measures = TRUE)
summary(mbasica_cfa, standardized = TRUE, fit.measures = TRUE)

modificationIndices(mmedia_cfa) %>% 
  filter(mi > 3.84)
```

```{r}
#| label: configural model
#| include: false 

baseline <- measEq.syntax(
  configural.model = model_restricted, # se especifica el modelo que se usará
  data = db1,
  ordered = c("perc_effort", "perc_talent", "perc_rich_parents", "perc_contact", # se explicita que las 
              "pref_effort", "pref_talent", "pref_rich_parents", "pref_contact"),# variables son ordinales 
  parameterization = "theta", #parametrización ideal para indicadores ordinales
  ID.fac = "std.lv", # identidica el factor con varianza 1
  ID.cat = "Wu.Estabrook.2016", # tipo de estrategia ideal para indicadores ordinales
  group = "cohort_level", # identificación de los grupos
  group.equal = "configural" # nivel de invarianza 
)

model.baseline <- as.character(baseline) # se crea un objeto con el modelo configural como caracter para que puedan visualizarse sus datos

fit.baseline <- cfa(model.baseline, data = db1, group = "cohort_level",
ordered = c("perc_effort", "perc_talent", "perc_rich_parents", "perc_contact", 
              "pref_effort", "pref_talent", "pref_rich_parents", "pref_contact"))

summary(fit.baseline, fit.measures = T, standardized = T)
modificationIndices(fit.baseline) %>% 
  filter(mi > 3.84)
```

```{r}
#| label: weak model
#| include: false

# los siguientes dos modelos siguen los pasos de análisis de Dubravka et al, 2019.

thresholds <- measEq.syntax(
configural.model = model_restricted,
data = db1,
ordered = c("perc_effort", "perc_talent", "perc_rich_parents", "perc_contact",
"pref_effort", "pref_talent", "pref_rich_parents", "pref_contact"),
parameterization = "theta",
ID.fac = "std.lv",
ID.cat = "Wu.Estabrook.2016",
group = "cohort_level",
group.equal = c("thresholds") # aquí se aplica lo mismo que el código anterior excepto que ahora se restringen los thresholds
)

model.thres <- as.character(thresholds)

fit.thres <- cfa(model.thres, data = db1, group = "cohort_level",
ordered = c("perc_effort", "perc_talent", "perc_rich_parents", "perc_contact",
"pref_effort", "pref_talent", "pref_rich_parents", "pref_contact"))

summary(fit.thres, fit.measures = TRUE)
lavTestScore(fit.thres)
pt <- lavaan::parTable(fit.thres)

pt[pt$plabel %in% c(".p9.", ".p10.", ".p80.", ".p81."),
   c("id","group","lhs","op","rhs","label","plabel","free","ustart","est")]
```

```{r}
#| label: loading-threshold model
#| include: false

strong <- measEq.syntax(
configural.model = model_restricted,
data = db1,
ordered = c("perc_effort", "perc_talent", "perc_rich_parents", "perc_contact",
"pref_effort", "pref_talent", "pref_rich_parents", "pref_contact"),
parameterization = "theta",
ID.fac = "std.lv",
ID.cat = "Wu.Estabrook.2016",
group = "cohort_level",
group.equal = c("thresholds", "loadings") # ahora se restringen los tresholds y las cargas factoriales
)

model.strong <- as.character(strong)

fit.strong <- cfa(model.strong, data = db1, group = "cohort_level",
ordered = c("perc_effort", "perc_talent", "perc_rich_parents", "perc_contact",
"pref_effort", "pref_talent", "pref_rich_parents", "pref_contact"))

summary(fit.strong, fit.measures = TRUE)

```

```{r}
#| echo: false
#| label: tbl-cohort2
#| tbl-cap: Cohort (multigroup) invariance results
#| tbl-cap-location: top

# Comparaciones de Chi²
an1 <- anova(fit.baseline, fit.thres)
an2 <- anova(fit.thres, fit.strong)

tab01 <- bind_rows(
  as_tibble(an1)[2,],
  as_tibble(an2)[2,] # aqui se toma la segunda fila de comparación de cada anova
) %>%
  select("Chisq", "Df", chisq.diff = `Chisq diff`, df.diff = `Df diff`, pvalue = `Pr(>Chisq)`) %>%
  mutate(
    stars = stars.pval(pvalue),
    chisqt = paste0(round(Chisq, 2), " (", Df, ")"),
    decision = ifelse(pvalue > 0.05, "Accept", "Reject"),
    model = c("Tresholds", "Tresholds + Loadings") # etiqueta los modelos que se extraen de los anova
  ) %>%
  bind_rows(
    tibble(
      Chisq = an1$Chisq[1],
      Df = an1$Df[1],
      chisq.diff = NA,
      df.diff = NA,
      pvalue = NA,
      stars = "",
      chisqt = paste0(round(an1$Chisq[1], 2), " (", an1$Df[1], ")"),
      decision = "Reference",
      model = "Configural" # se añade el modelo configural como referencia para clarificar la comparación
    )
  ) %>%
  select(model, chisqt, chisq.diff, df.diff, pvalue, stars, decision) %>%
  mutate(model = factor(model, levels = c("Configural", "Tresholds", "Tresholds + Loadings"))) %>%
  arrange(model)

# se extraen los índices de ajuste para cada modelo
fit.meas <- bind_rows(
  fitMeasures(fit.baseline, output = "matrix")[c("chisq.scaled", "df.scaled", "cfi.scaled", "rmsea.scaled", "rmsea.ci.lower.scaled", "rmsea.ci.upper.scaled"),],
  fitMeasures(fit.thres, output = "matrix")[c("chisq.scaled", "df.scaled", "cfi.scaled", "rmsea.scaled", "rmsea.ci.lower.scaled", "rmsea.ci.upper.scaled"),],
  fitMeasures(fit.strong, output = "matrix")[c("chisq.scaled", "df.scaled", "cfi.scaled", "rmsea.scaled", "rmsea.ci.lower.scaled", "rmsea.ci.upper.scaled"),]
)

fit.meas <- fit.meas %>% # se calculan las diferencias de cada modelo con el anterior para 
  rename_with(~str_remove(., ".scaled")) %>% 
  mutate(                # analizar si se cumplen criterios de invarianza 
    diff.chi2 = chisq - lag(chisq, default = first(chisq)),
    diff.df = df - lag(df, default = first(df)),
    diff.cfi = cfi - lag(cfi, default = first(cfi)),
    diff.rmsea = rmsea - lag(rmsea, default = first(rmsea))
  ) %>%
  round(3) %>%
  mutate(rmsea.ci = paste0(rmsea, " \n(", rmsea.ci.lower, "-", rmsea.ci.upper, ")"))

# Tabla final (mantener diff.df hasta formatear)
tab.inv <- bind_cols(tab01, fit.meas) %>%
  select(model, chisqt, cfi, rmsea.ci, diff.chi2, diff.df, diff.cfi, diff.rmsea, stars, decision) %>%
  mutate(diff.chi2 = paste0(diff.chi2, " (", diff.df, ") ", stars)) %>%
  select(model, chisqt, cfi, rmsea.ci, diff.chi2, diff.cfi, diff.rmsea, decision)

# Encabezados de columna
col.nam <- c(
  "Model",
  "$\\chi^2$ (df)",
  "CFI",
  "RMSEA (90\\% CI)",
  "$\\Delta \\chi^2$ ($\\Delta$df)",
  "$\\Delta$CFI",
  "$\\Delta$RMSEA",
  "Decision"
)
# Se formatea la tabla en kable
tab.inv %>%
  kableExtra::kable(
    format = "latex",
    align = "c",
    booktabs = TRUE,
    escape = F,
    col.names = col.nam
  ) %>%
  kableExtra::kable_styling(
    full_width = TRUE, 
    latex_options = "hold_position",
    bootstrap_options = c("striped", "bordered", "condensed"),
   font_size = 9.5) %>% 
  kableExtra::column_spec(column = 1, width = "2cm") %>%
   kableExtra::column_spec(column = c(4:5),width = "2 cm")

```


@tbl-cohort2 summarizes the cross-cohort invariance tests. The scale achieves configural invariance between primary and secondary students, but not threshold invariance, and therefore does not meet the prerequisites for stronger forms of invariance [@wu_identification_2016a; @svetina_multiplegroup_2020]. Constraining thresholds worsened model fit ($\Delta$CFI = −.013; $\Delta$RMSEA = .013), exceeding recommended cutoffs [@chen_sensitivity_2007]. This indicates that response categories do not map equivalently onto the latent constructs across cohorts, limiting direct comparisons of latent means or other latent differences.

To locate the source of cross-cohort non-invariance, we inspected localized misfit in the threshold-invariant model. The largest violations concerned the perceived-effort item, especially the second threshold ($\chi^2=10.12, p=.001$) and, to a lesser extent, the first ($\chi^2=7.55, p=.006$), suggesting that older students may apply a more critical standard when judging whether effort is rewarded in society.
